---
title: "Try sex inference"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 150
)
```

```{r, load libraries}
library(glmnet)
library(DESeq2)
library(tidyverse)
library(data.table)
library(caret)
library(glmnet)
library(here)
library(httpgd)
library(patchwork)

source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


## **Introduction**
We have downloaded the whole ARCH4S dataset. Using a python script (extract_mdata.ipynb) we have extracted
all samples in that dataset that come from macrophage and that have sex information available. 
Note: We have changed from macrophages to whole blood. Reason for this is that some "macrophage" datasets are 
acutally just using whole blood when you look them up...

Here, we will fit the lasso regression to predict sex.

```{r define functions}
inverse_rank_norm <- function(x) {
    qnorm((rank(x, na.last = "keep", ties.method="random") - 0.5) / sum(!is.na(x)))
}
```


```{r}
# load in mdata
mdata <- fread(here("sex_inference", "results", "mdata_whole_blood.tsv"))
mdata$sex <- ifelse(mdata$sex == "F", "FEMALE", "MALE")
mdata <- mdata %>% na.omit()

# only keep series ids that have used both males and females
keep_ids <- mdata %>%
    group_by(series_id, sex) %>%
    tally() %>%
    ungroup() %>%
    group_by(series_id) %>%
    tally() %>%
    filter(n == 2) %>%
    pull(series_id)

mdata <- mdata %>%
    filter(series_id %in% keep_ids)

# load in count data
data <- fread(here("sex_inference", "results", "counts_whole_blood.tsv.gz")) 

# only keep samples that pass mdata filter
data <- data.frame(data)[, c("gene", mdata$sampleID)]

# read in our data so we know which genes are "available"
our_dds <- readRDS(here("QC", "results", "dds_final.rds"))

# figure out which genes are shared between the datasets
shared_genes <- intersect(rownames(our_dds), data$gene)

data <- data %>%
    filter(gene %in% shared_genes)

# some genes occure multiple times... no idea why.
# we will just sum them together if this happens
data <- data %>%
    group_by(gene) %>%
    summarise(across(everything(), sum)) %>%
    column_to_rownames("gene") 

# reorder the gene names
data <- data[match(shared_genes, rownames(data)), ]

# create a deseq2 object for easy handling manipulation of the data
dds <- DESeqDataSetFromMatrix(
    countData = data,
    colData = mdata %>% column_to_rownames("sampleID"),
    design = ~1
)


#norm_data <- apply(counts(dds), 1, inverse_rank_norm)
vsd_data <- vst(dds, blind = TRUE)
norm_data <- assay(vsd_data)
```


### Some visualizations
```{r, eval=FALSE}
#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(norm_data,
    metadata = colData(dds),
    removeVar = 0.4
)

pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sampleID") %>%
    left_join(., mdata, by = c("sampleID" = "sampleID"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}


##################
# CALCULATE UMAP #
##################
umap_data <- umap::umap(t(norm_data))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleID") %>% 
    left_join(., mdata)


i = c(1, 2)
expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
pca_plot <- density_scatter(
    df = pca_plot_df,
    x_variable = "PC1",
    y_variable = "PC2",
    color_by = "series_id",
    density_plot_ratio = 0.15,
    pt_size = 1.2,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)

ggsave(
    plot=pca_plot,
    filename = here("sex_inference", "results", "PCA_unadjusted.png"),
    dpi=400,
    height=10,
    width=12
)


umap_plot <- density_scatter(
    df = umap_plot_df,
    x_variable = "UMAP1",
    y_variable = "UMAP2",
    color_by = "series_id",
    density_plot_ratio = 0.15,
    pt_size = 1.2,
)


```


### Adjust counts
```{r}
# adjust counts using limma
adjusted_counts <- limma::removeBatchEffect(
    norm_data,
    batch = dds$series_id,
    design = model.matrix(~sex, colData(dds))
)

# recalculate PCA with adjusted counts
pca_data_corrected <- PCAtools::pca(
    adjusted_counts,
    metadata = colData(dds),
    removeVar = 0.4
)

pca_plot_df <- pca_data_corrected$rotated[, 1:20] %>%
    rownames_to_column("sampleID") %>%
    left_join(., mdata, by = c("sampleID" = "sampleID"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data_corrected$variance[i]
}


##################
# CALCULATE UMAP #
##################
umap_data_corrected <- umap::umap(t(adjusted_counts))
umap_plot_df <- umap_data_corrected$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleID") %>% 
    left_join(., mdata)


i = c(1, 2)
expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)

pca_plot_corrected <- density_scatter(
    df = pca_plot_df,
    x_variable = "PC1",
    y_variable = "PC2",
    color_by = "series_id",
    density_plot_ratio = 0.15,
    pt_size = 1.2,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)


umap_plot_corrected <- density_scatter(
    df = umap_plot_df,
    x_variable = "UMAP1",
    y_variable = "UMAP2",
    color_by = "series_id",
    density_plot_ratio = 0.15,
    pt_size = 1.2,
)


ggsave(
    plot = pca_plot_corrected,
    filename = here("sex_inference", "results", "PCA_corrected.png"),
    dpi = 400,
    height = 10,
    width = 12
)

```


## Fit the model

```{r}
mdata <- fread(here("sex_inference", "results", "mdata_whole_blood.tsv"))
mdata$sex <- ifelse(mdata$sex == "F", "FEMALE", "MALE")
mdata <- mdata %>% na.omit()

# only keep series ids that have used both males and females
keep_ids <- mdata %>%
    group_by(series_id, sex) %>%
    tally() %>%
    ungroup() %>%
    group_by(series_id) %>%
    tally() %>%
    filter(n == 2) %>%
    pull(series_id)

mdata <- mdata %>%
    filter(series_id %in% keep_ids)

# Split the data into training and test set
data <- t(adjusted_counts)


#data <- data[, c("XIST", predictive_genes)]


set.seed(50)
training_samples <- mdata$sex %>% 
  createDataPartition(p = 0.85, list = FALSE)
train_data  <- data[training_samples, ]
test_data <- data[-training_samples, ]
train_mdata <- mdata[training_samples, ]
test_mdata <- mdata[-training_samples, ]


# Prepare the matrices for glmnet
x_train <- as.matrix(train_data)
y_train <- train_mdata$sex

x_test <-  as.matrix(test_data)
y_test <- test_mdata$sex


# Fit logistic regression using cross validation

cv_fit <- cv.glmnet(x_train, y_train,
    family = "binomial",
    type.measure = "class",
    n.folds = 10,
    alpha = 1 # 1 corresponds to lasso regression
)
plot(cv_fit)

# Best lambda
best_lambda <- cv_fit$lambda.min
print(best_lambda)


# Final model with lambda.min
# this uses all of the train data
# the cv model above just the best fold, i.e. misses 1/10 of the data
lasso.model <- glmnet(x_train, y_train,
    alpha = 1, family = "binomial",
    lambda = best_lambda
)


saveRDS(
    lasso.model,
    here("sex_inference", "results", "lasso_model_sex.rds")
)


predictive_genes <- coef(lasso.model) %>%
    as.matrix() %>%
    data.frame() %>%
    filter(s0 != 0) %>%
    arrange(s0) %>%
    rownames()


# Make prediction on test data
probabilities <- lasso.model %>% predict(
    newx = x_test,
    type = "response"
)
predicted.classes <- ifelse(probabilities > 0.5, "MALE", "FEMALE")

# Model accuracy
mean(predicted.classes == y_test)
```


### Plot the prediction
```{r}
dds <- estimateSizeFactors(dds)
plot_data_train <- t(log1p(counts(dds, normalize=TRUE)))[,predictive_genes[-which(predictive_genes == "(Intercept)")]] %>%
    data.frame() %>%
    rownames_to_column("sampleid") %>%
    pivot_longer(., cols = !sampleid, values_to = "norm_cnts", names_to = "gene") %>%
    left_join(., mdata %>% select(sampleid=sampleID, sex), by = "sampleid")

p <- ggplot(plot_data_train, aes(x = sex, y = norm_cnts)) +
    ggbeeswarm::geom_quasirandom(width = 0.15, size = 0.1) +
    facet_wrap(~gene, scales = "free_y") +
    theme_Publication() +
    stat_summary(
        fun = "mean",
        geom = "crossbar",
        color = "salmon",
        width = 0.1,
        aes(group = sex),
        position = position_dodge(width = 0.5)
    )

ggsave(
    plot=p,
    filename = here("sex_inference", "results", "training_boxplots.png"),
    width = 7,
    height=5,
    dpi=350,
    scale=1.35
)
```




### Run model on our data

```{r}
library(sva)
dds <- readRDS(here("QC", "results", "dds_final.rds"))
# order the genes like our training data
dds <- dds[rownames(dds) %in% shared_genes, ]
dds <- dds[match(shared_genes, rownames(dds)),]
stopifnot(all(colnames(data) == rownames(dds)))

norm_data <- assay(vst(dds, blind = TRUE))

# runs long, even with parallization
# mod <- model.matrix(~ line + group + sex, colData(dds))
# mod0 <- model.matrix(~ line + group, colData(dds))
# svseq <- svaseq(norm_data, mod, mod0, n.sv = 5)
# saveRDS(svseq, here("QC", "results", "sva_5_final_sex_compatibility.rds"))

svseq <- readRDS(here("QC", "results", "sva_5_final_sex_compatibility.rds"))
SVs <- svseq$sv
colnames(SVs) <- paste0("SV", 1:5)
mdata <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)
mod <- model.matrix(~sex, mdata)
# Create the formula dynamically
formula <- as.formula(paste("~", paste0("SV", 1:5, collapse = " + ")))
mod0 <- model.matrix(formula, mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

# adjust the counts using limma
adjusted_counts_our_data <- limma::removeBatchEffect(norm_data,
    batch = dds$line,
    batch2 = dds$group,
    covariates = mod0,
    design = mod
) %>% t()



stopifnot(all(colnames(adjusted_counts_our_data) == colnames(data)))
probabilities_our_data <- lasso.model %>% predict(
    newx = adjusted_counts_our_data,
    type = "response"
)
our_predicted_class <- ifelse(probabilities_our_data > 0.5, "Male", "Female")

all(mdata$sampleid == rownames(our_predicted_class))

mdata$predicted_sex <- our_predicted_class[,1]

table(mdata$predicted_sex != mdata$sex)
```

```{r}
# for each celline, calculate the average expression of these genes for all conditions
plot_data <- t(log1p(counts(dds, normalize=TRUE)))[,predictive_genes[-which(predictive_genes == "(Intercept)")]] %>%
    t() %>%
    data.frame() %>%
    rownames_to_column("sampleid") %>%
    pivot_longer(., cols = !sampleid, values_to = "norm_cnts", names_to = "gene") %>%
    left_join(., mdata %>% select(sampleid, line, sex, predicted_sex), by = "sampleid") %>%
    group_by(line, gene) %>%
    mutate(mean_expr = mean(norm_cnts)) %>%
    ungroup() %>%
    distinct(gene, line, mean_expr, .keep_all = TRUE)

plot_data$mismatch <- ifelse(plot_data$sex != plot_data$predicted_sex, "mismatch", "match")

ggplot(plot_data, aes(x = sex, y = mean_expr, color = mismatch)) +
    geom_jitter(width = 0.1) +
    facet_wrap(~gene, scales = "free_y") +
    theme_Publication() +
    stat_summary(
        fun = "mean",
        geom = "crossbar",
        color = "darkgreen",
        width = 0.05,
        aes(group = sex),
        position = position_dodge(width = 0.5)
    ) +
    scale_color_manual(values = c("black", "red")) +
    theme(legend.position = "none")

```