---
title: "RNA inital QC"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 150
)
```

## **Introduction**
The [nf-core RNAseq pipeline](https://github.com/nf-core/rnaseq) already does a very extensive set of quality control steps 
based on the read level (for example # Duplicated reads, GC content, 
mapping rate, read distribution, …). 
In this script, we will do some qulaity control steps based on the 
quantified counts (using [STAR](https://github.com/alexdobin/STAR) followed by [salmon](https://combine-lab.github.io/salmon/)).
 We will also make a few plots to explore the metadata


```{r, load libraries}
library(DESeq2)
library(magrittr)
library(PCAtools)
library(tidyverse)
library(umap)
library(patchwork)
library(httpgd)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


## **Description of metadata**
Few observations: 
(a) A few important confounders (e.g. `age`, `ethnicity`) are missing for some samples 
(b) There seem to be a few technical replicates (cell lines that were used more than 6 times) 
(c) MM_oxLDL7159504 seems to contain conflicting information. 
oxLDL treated according to column `oxLDL` but control according to column `sampeltype`.

```{r}
mdata <- readxl::read_excel(here("data", "mdata_google_spreadsheet.xlsx"),
    sheet = "RNA Sequencing samples"
) %>% data.frame()

submission_mdata <- data.table::fread(here("data", "oxLDL_data", "RNA_seq", "macromap_samplesheet.csv"),
    data.table = FALSE
)

# reformat some mdata entries
mdata <- mdata %>%
    dplyr::rename(condition = oxldl) %>%
    dplyr::mutate(condition = ifelse(condition == 100, "stim", "ctrl")) %>%
    dplyr::rename(sex = set) %>%
    dplyr::mutate(group = paste(state, condition, sep = "_")) %>%
    dplyr::select(sampleid, line, group, state, condition, sex, ethnicity, age, oxldlset, pool, batch1id)

mdata$ethnicity <- case_when(
    mdata$ethnicity == "White - White British" ~ "White British",
    mdata$ethnicity == "White - White other" ~ "White Other",
    mdata$ethnicity == "White - Other" ~ "White Other",
    mdata$ethnicity == "Asian /Asian British- Indian" ~ "Asian British",
    T ~ mdata$ethnicity
)

# define some factors
mdata <- mdata %>%
    mutate(
        group = factor(group, levels = c("M0_ctrl", "M1_ctrl", "M2_ctrl", "M0_stim", "M1_stim", "M2_stim")),
        line = factor(line),
        state = factor(state, levels = c("M0", "M1", "M2")),
        condition = factor(condition, levels = c("ctrl", "stim")),
        sex = factor(sex, levels = c("Female", "Male")),
        ethnicity = factor(ethnicity, levels = c("White British", "White Other", "Asian British", "NA")),
        oxldlset = factor(oxldlset, levels = sort(unique(mdata$oxldlset))),
        age = factor(age, levels = sort(unique(mdata$age))),
        pool = factor(pool, levels = str_sort(unique(mdata$pool), numeric=T)),
        batch1id = factor(batch1id, levels = str_sort(unique(mdata$batch1id), numeric=T)),
    )

# make sure all samples ocurre in the submission script
stopifnot(all(mdata$sampleid %in% submission_mdata$sample))
stopifnot(n_distinct(mdata$sampleid) == nrow(mdata))

age <- plot_column_table(
    mdata %>%
        distinct(line, .keep_all = T) %>%
        pull(age),
    include_n = T
) +
    ggtitle("Age distribution") +
    xlab("Age") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))


rna_pool <- plot_column_table(mdata$pool %>%
    str_remove_all("RNA ") %>%
    factor(., levels = c("pilot", as.character(1:17))), include_n=TRUE) +
    ggtitle("RNA pool distribution") +
    xlab("RNA pool")


state <- plot_column_table(mdata$state, include_n = T) +
    ggtitle("Macrophage group distribution") +
    xlab("Macrophage group")

# some lines have more than the expected number (6) of entries. Take a look at them
# we might have to remove one of them because they are technical replicates and not biological ones.
lines <- plot_column_table(mdata$line) +
    ggtitle("Cell line distribution") +
    xlab("Cell lines") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

# simple technical replicates sometimes, even in the same RNA pool and batch
# for example MM_oxLDL9752052 and MM_oxLDL9752058
mdata %>%
    group_by(line) %>%
    mutate(n_lines = n()) %>%
    filter(n_lines > 6) %>%
    arrange(line) 

condition <- plot_column_table(mdata$condition,
    include_n = T
) +
    ggtitle("Condition distribution") +
    xlab("Condition")


oxLDL_set <- plot_column_table(factor(mdata$oxldlset))+
    ggtitle("oxLDL set distribution") +
    xlab("oxLDL set") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

sex <- plot_column_table(
    mdata %>%
        distinct(line, .keep_all = T) %>%
        pull(sex),
    include_n = T
) +
    ggtitle("Sex distribution") +
    xlab("Sex")

ethnicity <- plot_column_table(mdata$ethnicity,
    include_n = TRUE
) +
    ggtitle("Ethnicity") +
    xlab("Ethnicity") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

batches <- plot_column_table(as.character(mdata$batch1id), include_n = TRUE)+
    ggtitle("Batches distribution") +
    xlab("Batch")+
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

    # combine the plots
combined_plot <- wrap_plots(
    age, sex, ethnicity,
    lines, oxLDL_set, batches,
    state, condition,
    ncol = 3,
    byrow = TRUE
)
ggsave(combined_plot,
    filename = here("QC", "results", "imgs", "mdata_overview.png"),
    dpi = 150, width = 18, height = 8
)
data.table::fwrite(mdata, here("QC", "results", "mdata_clean.tsv"), sep="\t", row.names=FALSE)
# so factor information remains
saveRDS(mdata, here("QC", "results", "mdata_clean.rds"))
```

## **Count data quality control**
Read in the count data generated by the pipeline. They were slightly modified using the 
script preprocessing/scripts/unify_geneID_to_geneName.rmd


```{r}
raw_counts <- data.table::fread(
    here("preprocessing", "results", "gene_counts_length_scaled_filtered.tsv")
) %>% data.frame()
raw_counts_matrix <- as.matrix(raw_counts[,2:ncol(raw_counts)])
rownames(raw_counts_matrix) <- raw_counts$gene_name

# make sure mdata order matches the column names of raw_counts
mdata <- mdata[match(colnames(raw_counts_matrix), mdata$sampleid), ]
libsize <- colSums(round(raw_counts_matrix)) %>% enframe(name="sampleid", value="libsize")
mdata <- mdata %>% left_join(.,libsize, by="sampleid")
rownames(mdata) <- mdata$sampleid
data.table::fwrite(mdata, here("QC", "results", "mdata_clean.tsv"), sep="\t", row.names=FALSE)
# so factor information remains
saveRDS(mdata, here("QC", "results", "mdata_clean.rds"))
dds0 <- DESeqDataSetFromMatrix(round(raw_counts_matrix),
    colData = mdata,
    design = ~1
)

# remove some lowly expressed genes
keep <- edgeR::filterByExpr(counts(dds0), group = mdata$group)
dds0 <- dds0[keep,]
vsd_data <- vst(dds0)
norm_data <- assay(vsd_data)
saveRDS(dds0, here("QC", "results", "dds_original.rds"))
```

### Create QC plots for uncorrected data
Clearly strong batch effects. Main culprits are:
(a) batch1id
(b) pool (overlaps with batch1id)
(c) age
Our variables of interest seem to not associate with any PCs.

Furthermore, we observe 7 samples which are far away from the rest. Further investigate these.
```{r}
#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(norm_data,
    metadata = colData(dds0),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

##################
# CALCULATE UMAP #
##################
umap_data <- umap(t(norm_data))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>% 
    left_join(., mdata)

###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
    components = paste0("PC", 1:15)
)

png(here("QC", "results", "imgs", "all_samples","association_heatmap.png"), res=150, width=9, height=5.5, units="in")
association_output$heatmap
dev.off()

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")

PC_pairings <- lapply(1:7, function(x){
    c(2*x-1,2*x)
})

for (i in PC_pairings) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "QC", "results", "imgs", "all_samples", str_interp("PCA_all_covariates_PC${i[1]}_PC${i[2]}.png")
        ),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
}


################
# BOXPLOTS PCA #
################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = pca_plot_df,
        covar = x,
        PC = paste0("PC", 1:4)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "all_samples", "PCA_all_covariates_boxplots.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)


##############
# UMAP PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "all_samples", "UMAP_all_covariates.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)

################
# BOXPLOTS UMAP #
################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = umap_plot_df,
        covar = x,
        PC = paste0("UMAP", 1:2)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "all_samples", "UMAP_all_covariates_boxplots.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)
```

### Identify and justify outliers
For PC1 and PC2 there are some huge outliers that drive the majority of variation in them. Identify them and try to see 
why they behave so differently. The outliers have PC1 > "stim" and PC2 < -"stim"

```{r}
outlier_samples <- pca_plot_df %>%
    filter(PC1 > 100) %>%
    filter(PC2 < - 100) %>%
    pull(sample_id)

########### 
# libsize #
###########
low_libsize <- mdata %>% slice_min(., order_by = libsize, n=7) %>% pull(sampleid)

# all have extremely low library size
all(outlier_samples %in% low_libsize)

mdata_small_lib <- mdata %>%
    slice_min(., order_by = libsize, n = 100) %>%
    arrange(libsize) %>%
    mutate(sorted_index = row_number())

ggplot(
    mdata_small_lib, aes(x = sorted_index, y = libsize)) +
    geom_bar(stat = "identity") +
    labs(x = "Sorted library size index", y = "Library Size") +
    theme_Publication() +
    scale_y_continuous(expand = c(0.01,0))


libsizes <- sort(mdata$libsize)
ratios <- libsizes[-1] / libsizes[-length(libsizes)]


# Create a data frame for plotting
df <- data.frame(index = 1:(length(libsizes)-1), ratio = ratios)

# Plot the ratios
ggplot(df, aes(x = index, y = ratio)) +
    geom_point() +
    labs(x = "Index", y = "(n+1)/n Ratio") +
    ggtitle("Ratios of (n+1)/n for sorted library size") +
    theme_Publication()

ggplot(pca_plot_df, aes(PC1, PC2)) +
    geom_point() +
    geom_point(data = pca_plot_df %>% filter(sample_id == "MM_oxLDL8032568"),
        aes(PC1, PC2, color=sample_id), inherit.aes = FALSE) +
    theme_Publication()

ggplot(pca_plot_df, aes(PC1, PC2)) +
    geom_point() +
    geom_point(
        data = pca_plot_df %>%
            filter(libsize < 5e6),
        aes(PC1, PC2), inherit.aes = FALSE, size = 2, color="steelblue"
    ) +
    theme_Publication() %+%
    theme(legend.position = "none")


ggplot(pca_plot_df, aes(PC1, PC2, color=log(libsize))) +
    geom_point() 


ggplot(pca_plot_df, aes(PC1, libsize)) +
    geom_point() +
    geom_smooth(method = "lm")



# for PC3 and 4 it seems like the samples below 5e6 are making a big difference
i = c(1, 2)
density_scatter(
    df = pca_plot_df %>% mutate(low_libsize_dummy = ifelse(libsize < 1e4, "low libsize", "normal libsize")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "low_libsize_dummy",
    density_plot_ratio = 0.15,
    pt_size = 4,
    xlab = str_interp("PC${i[1]}"),
    ylab = str_interp("PC${i[2]}")
)


```


### Remove outliers and rerun PCA/UMAP
Now we do not have any clear outliers, however we still see strong batch effects. 
In the next Rmd (02_batch_correction.rmd), we will try different methods to currect for that. 

```{r}
outlier_samples <- pca_plot_df %>%
    filter(PC1 > 100) %>%
    filter(PC2 < -100) %>%
    pull(sample_id)

dds0_filtered <- dds0[, !dds0$sampleid %in% outlier_samples]
saveRDS(dds0_filtered, here("QC", "results", "dds_libsize_outliers_removed.rds"))
vsd_data <- vst(dds0_filtered)
norm_data <- assay(vsd_data)

#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(norm_data,
    metadata = colData(dds0_filtered),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

##################
# CALCULATE UMAP #
##################
umap_data <- umap(t(norm_data))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>% 
    left_join(., mdata)

###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
    components = paste0("PC", 1:15)
)

png(here("QC", "results", "imgs", "libsize_outliers_removed","association_heatmap.png"), res=150, width=9, height=5.5, units="in")
association_output$heatmap
dev.off()

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")

PC_pairings <- lapply(1:7, function(x){
    c(2*x-1,2*x)
})
for (i in PC_pairings) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )


        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "QC", "results", "imgs", "libsize_outliers_removed",str_interp("PCA_all_covariates_PC${i[1]}_PC${i[2]}.png")
        ),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
}

################
# BOXPLOTS PCA #
################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = pca_plot_df,
        covar = x,
        PC = paste0("PC", 1:4)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "libsize_outliers_removed", "PCA_all_covariates_boxplots.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)


##############
# UMAP PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "libsize_outliers_removed", "UMAP_all_covariates.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)

#################
# BOXPLOTS UMAP #
#################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = umap_plot_df,
        covar = x,
        PC = paste0("UMAP", 1:2)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "libsize_outliers_removed", "UMAP_all_covariates_boxplots.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)


# we can still see an effect for samples with low number of total reads. 
i = c(1, 2)
i = c(3, 4)
density_scatter(
    df = pca_plot_df %>% mutate(low_libsize_dummy = ifelse(libsize < 5e6, "low libsize", "normal libsize")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "low_libsize_dummy",
    density_plot_ratio = 0.15,
    pt_size = 4,
    xlab = str_interp("PC${i[1]}"),
    ylab = str_interp("PC${i[2]}"),
)
```


### Remove samples with less than 5million reads
We will remove all samples with less than 5 million reads. 
This is the general guideline (for example for ENCODE) and we can see that samples with less than 
5 million reads cluster together on PC3 and PC4
In the next Rmd (02_batch_correction.rmd), we will try different methods to currect for that. 

```{r}
outlier_samples <- colData(dds0) %>%
    data.frame() %>%
    filter(libsize < 5e6) %>%
    pull(sampleid)

dds0_filtered <- dds0[, !dds0$sampleid %in% outlier_samples]
saveRDS(dds0_filtered, here("QC", "results", "dds_libsize_5e6_removed.rds"))
vsd_data <- vst(dds0_filtered)
norm_data <- assay(vsd_data)

#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(norm_data,
    metadata = colData(dds0_filtered),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

##################
# CALCULATE UMAP #
##################
umap_data <- umap(t(norm_data))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>% 
    left_join(., mdata)

###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
    components = paste0("PC", 1:15)
)

png(here("QC", "results", "imgs", "libsize_5e6_removed","association_heatmap.png"), res=150, width=9, height=5.5, units="in")
association_output$heatmap
dev.off()

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")

PC_pairings <- lapply(1:7, function(x){
    c(2*x-1,2*x)
})
for (i in PC_pairings) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )


        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "QC", "results", "imgs", "libsize_5e6_removed",str_interp("PCA_all_covariates_PC${i[1]}_PC${i[2]}.png")
        ),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
}

################
# BOXPLOTS PCA #
################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = pca_plot_df,
        covar = x,
        PC = paste0("PC", 1:4)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "libsize_5e6_removed", "PCA_all_covariates_boxplots.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)

##############
# UMAP PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "libsize_5e6_removed", "UMAP_all_covariates.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)

#################
# BOXPLOTS UMAP #
#################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = umap_plot_df,
        covar = x,
        PC = paste0("UMAP", 1:2)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "libsize_5e6_removed", "UMAP_all_covariates_boxplots.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)
```

### Updated mdata
Check how the mdata changed after removing the samples

```{r, eval=F}
mdata <- colData(dds0_filtered) %>% data.frame()
age <- plot_column_table(
    mdata %>%
        distinct(line, .keep_all = T) %>%
        pull(age),
    include_n = T
) +
    ggtitle("Age distribution") +
    xlab("Age") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))


rna_pool <- plot_column_table(mdata$pool %>%
    str_remove_all("RNA ") %>%
    factor(., levels = c("pilot", as.character(1:17))), include_n=TRUE) +
    ggtitle("RNA pool distribution") +
    xlab("RNA pool")


group <- plot_column_table(mdata$state, include_n = T) +
    ggtitle("Macrophage group distribution") +
    xlab("Macrophage group")

# some lines have more than the expected number (6) of entries. Take a look at them
# we might have to remove one of them because they are technical replicates and not biological ones.
lines <- plot_column_table(mdata$line) +
    ggtitle("Cell line distribution") +
    xlab("Cell lines") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

# simple technical replicates sometimes, even in the same RNA pool and batch
# for example MM_oxLDL9752052 and MM_oxLDL9752058
mdata %>%
    group_by(line) %>%
    mutate(n_lines = n()) %>%
    filter(n_lines > 6) %>%
    arrange(line) 

condition <- plot_column_table(ifelse(mdata$condition == ""stim"", "stimulated", "unstimulated"),
    include_n = T
)+
    ggtitle("Condition distribution") +
    xlab("Condition")


oxLDL_set <- plot_column_table(factor(mdata$oxldlset))+
    ggtitle("oxLDL set distribution") +
    xlab("oxLDL set") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

sex <- plot_column_table(
    mdata %>%
        distinct(line, .keep_all = T) %>%
        pull(set),
    include_n = T
) +
    ggtitle("Sex distribution") +
    xlab("Sex")

ethnicity <- plot_column_table(
    factor(ethnicity_vector <- mdata %>%
    distinct(line, .keep_all = TRUE) %>%
    pull(ethnicity),
        levels = c("White - White British", "White - Other", "Asian British- Indian", "NA")
    ),
    include_n = TRUE
) +
    ggtitle("Ethnicity") +
    xlab("Ethnicity") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

batches <- plot_column_table(as.character(mdata$batch1id), include_n = TRUE)+
    ggtitle("Batches distribution") +
    xlab("Batch")+
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

    # combine the plots
combined_plot <- wrap_plots(
    age, sex, ethnicity,
    lines, oxLDL_set, batches,
    group, condition,
    ncol = 3,
    byrow = TRUE
)
ggsave(combined_plot,
    filename = here("QC", "results", "imgs", "mdata_overview_libsize_5e6_filtered.png"), 
    dpi = 150, width = 18, height = 8
)
```


### ENCODE QC recommendations
https://www.encodeproject.org/data-standards/rna-seq/long-rnas/
"Replicate concordance: the gene level quantification should have a Spearman correlation of >0.9 between
isogenic replicates and >0.8 between anisogenic replicates (i.e. replicates from different donors)."


```{r, eval=F}
dds <- readRDS(here("QC", "results", "dds_libsize_5e6_removed.rds"))
dds <- estimateSizeFactors(dds)
mdata <- colData(dds) %>% data.frame()
norm_cnts <- counts(dds, normalize=TRUE)

# we do have some isogenic replicates:
technical_repl_lines <- names(table(mdata$line))[table(mdata$line) > 6]


technical_replicate_samples <- mdata %>%
    group_by(samplename) %>%
    mutate(n = n()) %>%
    filter(n == 2) %>%
    arrange(samplename) %>% 
    pull(sampleid)

pairings <- lapply(1:(length(technical_replicate_samples)/2), function(x) {
    c(2 * x - 1, 2 * x)
})

pairwise_cor <- sapply(pairings, function(i){
    tmp_samples <- technical_replicate_samples[i]
    cor(norm_cnts[, tmp_samples[1]], norm_cnts[, tmp_samples[2]], method="spearman")
})
table(pairwise_cor > 0.9)



# anisogenic replicates
mdata <- mdata %>%
    mutate(group = paste(state, condition, sep="_")) 

cor_result <- lapply(unique(mdata$group), function(x){
    samples <- mdata %>%
        filter(group == x)  %>% 
        pull(sampleid)

    tmp_cnts <- norm_cnts[, samples]
    cor_map <- cor(tmp_cnts, method = "spearman")
    return(cor_map)
})
names(cor_result) <- unique(mdata$group)

table(cor_result$M0_0 > 0.8)

dim(cor_result$M0_0 )
library(ComplexHeatmap)
Heatmap(cor_result$M0_0) %>% ggplotify::as.ggplot()

```

```{r}
sessionInfo()
```
