---
title: "RNA seq batch correction"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 150
)
```

## **Introduction**
We saw in the script **01_initial_QC.rmd**, that there are strong batch effects in our dataset which is not surprising.
In this script, we will try out and evaluate different methods to correct for that. 

Here we adjust the data where we have removed samples with extremely low counts < 10k.

We noticed, a few things by now: 
(a) if we just use SVA, it looks like there is still residual batch effects for batch1id, line and others.
(b) if we instead adjust using "line" from the mdata, this looks much better, but PC1 and PC2 actually dont correspond to the biological process of interest. 
This means that there is, by definition, a lot of unwanted variation. 

Now we will try to combine these approaches. Adjust for line but also use SVA to model the unadjusted batch. 

Not 100% sure what the correct way for this is. I am going of the following forum posts:
https://support.bioconductor.org/p/54719/
https://support.bioconductor.org/p/98461/ -> see comment from asaffa01
https://support.bioconductor.org/p/124979/ -> this actually very useful if I decide to go with limma
https://www.biostars.org/p/333597/

```{r, load libraries}
library(DESeq2)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(magrittr)
library(PCAtools)
library(umap)
library(patchwork)
library(sva)
library(clusterProfiler)
library(RUVSeq)
library(httpgd)
library(tidyverse)
library(Rtsne)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```

```{r}
calc_pairwise_cor <- function(mat, outliers, condition_){
    keep_samples <- mdata %>% dplyr::filter(group == condition_) %>% pull(sampleid)
    tmp_cnts <- mat[, keep_samples]
    tmp_cnts <- tmp_cnts[, !colnames(tmp_cnts) %in% outliers]
    pairwise_correlation <- cor(tmp_cnts, method = "spearman")
    h <- Heatmap(pairwise_correlation,
        show_row_names = F,
        show_column_names = F,
        show_row_dend = F,
        show_column_dend = F,
        col = col_fun,
        name="R"
    ) %>%
        ggplotify::as.ggplot()
    return(list(
        heatmap = h,
        correlation_matrix = pairwise_correlation
    ))
}
```

## **SVA**

```{r calculate SVs}
# load in DESeq2dataset where the libsize outliers are already removed.
# for creation see script **01_initial_QC.rmd**
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
dds <- estimateSizeFactors(dds)
norm_counts <- counts(dds, norm=TRUE)
mod <- model.matrix(~line+group, colData(dds))
mod0 <- model.matrix(~line, colData(dds))



# # runs long, even with parallization
# dummy_ <- parallel::mclapply(3:20, function(i) {
#     svseq <- svaseq(norm_counts, mod, mod0, n.sv = i)
#     saveRDS(svseq, here("QC", "results", str_interp("sva_incl_line_${i}SVs.rds")))
# }, mc.cores=20)


# # alternative is this. I dont think this is correct though. actually causes an error: system is computationally singular
# mod <- model.matrix(~line+group, colData(dds))
# mod0 <- model.matrix(~1, colData(dds))
# dummy_ <- parallel::mclapply(3:20, function(i) {
#     svseq <- svaseq(norm_counts, mod, mod0, n.sv = i)
#     saveRDS(svseq, here("QC", "results", str_interp("sva_incl_line_null_model${i}SVs.rds")))
# }, mc.cores=20)


```


### **Try and find out what the SVs correspond to**
Still difficult to exactly determine what SV1 corresponds to. SV2 very clearly batch, but that was already the case for the PCA.
Big difference is that none of the SVs is associated with our conditions of interest which is good.
```{r}
mdata <- colData(dds) %>% data.frame()
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")

dummy_ <- parallel::mclapply(3:20, function(i) {
    dir.create(here("QC", "results", "imgs", str_interp("SVA_incl_line${i}")), showWarnings = FALSE)
    svseq <- readRDS(here("QC", "results", str_interp("sva_incl_line_${i}SVs.rds")))
    SVs <- svseq$sv
    colnames(SVs) <- paste0("SV", 1:i)
    mdata <- cbind(mdata %>% select(-contains("SV")), data.frame(SVs))

    ###########
    # HEATMAP #
    ###########
    covar_list = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize")
    correlation_SVs <- covar_correlation(
        df = mdata,
        covars = covar_list,
        components = paste0("SV", 1:i)
    )
    png(here("QC", "results", "imgs", str_interp("SVA_incl_line${i}"), "heatmap.png"), res = 150, width = 9, height = 5.5, units = "in")
    correlation_SVs$heatmap
    dev.off()

    ############
    # BOXPLOTS #
    ############
    # check for association
    for (x in columns_to_plot) {
        print(x)
        p <- plot_categorical_covar(
            df = mdata,
            covar = x,
            PC = paste0("SV", 1:i),
            bonferroni_correction = TRUE
        )
        p <- lapply(p, function(k) {
            k <- k + geom_hline(yintercept = 0, linetype = "dashed")
        })

        p <- p %>% wrap_plots(., ncol = 5) &
            theme(
                axis.text.x = element_blank(),
                axis.text.y = element_text(size = 5, face = "bold")
            )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        ggsave(
            plot = p,
            filename = here("QC", "results", "imgs", str_interp("SVA_incl_line${i}"), str_interp("SV_associations_${x}.png")),
            dpi = 150,
            width = 15,
            height = 10,
        )
    }

    ############
    # SV PLOTS #
    ############
    # similar to PC plots but with SV on the axis. See if this improves seperation of batches
    SV_pairings <- lapply(1:floor(i / 2), function(x) {
        c(2 * x - 1, 2 * x)
    })

    for (k in SV_pairings) {
        print(str_interp("Working on SV${k[1]} and SV${k[2]}"))
        plots <- lapply(columns_to_plot, function(x) {
            p <- density_scatter(
                df = mdata,
                x_variable = str_interp("SV${k[1]}"),
                y_variable = str_interp("SV${k[2]}"),
                color_by = x,
                density_plot_ratio = 0.15,
                pt_size = 1.2
            )


            p <- p +
                plot_annotation(
                    title = x,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
                )
            return(p)
        })
        final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
        ggsave(
            plot = final_plot,
            filename = here(
                "QC", "results", "imgs", str_interp("SVA_incl_line${i}"),
                str_interp("SVA_all_covariates_SV${k[1]}_SV${k[2]}.png")
            ),
            dpi = 150,
            width = 18,
            height = 11.5,
        )
    }
}, mc.cores = 20)

```


### Use SVs to correct the count data
```{r}
# this is needed for the enrichment analysis that we are doing.
hs_msigdb_df <- msigdbr::msigdbr(species = "Homo sapiens")

gene_sets <- hs_msigdb_df %>%
    filter(gs_subcat == "GO:BP" | gs_cat == "H" | gs_subcat == "CP:KEGG") %>%
    select(gs_name, gene_symbol)

vsd_data <- vst(dds, blind=TRUE)
mdata <- colData(dds) %>% data.frame()

dummy_ <- parallel::mclapply(3:20, function(i){
    dir.create(here("QC", "results", "imgs", str_interp("SVA_incl_line${i}")), showWarnings = FALSE)
    svseq <- readRDS(here("QC", "results", str_interp("sva_incl_line_${i}SVs.rds")))
    SVs <- svseq$sv
    colnames(SVs) <- paste0("SV", 1:i)
    tmp_mdata <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)

    # use limmas removeBatchEffects to emulate what the linear model will eventually do.
    norm_data <- assay(vsd_data)
    mod <- model.matrix(~group, tmp_mdata)
    # Create the formula dynamically
    formula <- as.formula(paste("~", paste0("SV", 1:i, collapse = " + ")))
    mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

    # adjust the counts using limma
    adjusted_counts <- limma::removeBatchEffect(norm_data,
        batch = vsd_data$line,
        covariates = mod0,
        design = mod
    )
    # recalculate PCA with adjusted counts
    pca_data <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds),
        removeVar = 0.4
    )

    # Extract the genes which most influence PC1 and PC2 and see if they are enriched in any biological domain:
    for (z in 1:2) {
        print(paste("Running ORA for PC", z))
        # extract top 300 genes which most influence PC
        top_loadings_gene_names <- pca_data$xvars[order(abs(pca_data$loadings[, paste0("PC", z)]), decreasing = T)[1:300]]

        enrichment_result <- enricher(
            gene = top_loadings_gene_names,
            universe = rownames(dds),
            minGSSize = 5,
            maxGSSize = 300,
            TERM2GENE = gene_sets
        )

        # for each of the biological domains, grep the top 3 sig. terms
        # for each of the biological domains, grep the top 3 sig. terms
        sig_results <- enrichment_result@result %>%
            mutate(domain = str_extract(Description, "^[^_]*")) %>%
            mutate(ratio_numeric = sapply(.$GeneRatio, function(x) {
                eval(
                    str2expression(x)
                )
            })) %>%
            filter(p.adjust < 0.05) %>%
            group_by(domain) %>%
            slice_min(., order_by = p.adjust, n = 3) %>%
            mutate(Description_short = shorten_strings(Description, 30))
        
        if (nrow(sig_results) > 0) {
            enrichment_plot <- ggplot(sig_results, aes(x = ratio_numeric, y = Description_short, color = p.adjust)) +
                geom_segment(aes(
                    x = 0, xend = ratio_numeric, y = Description_short, yend = Description_short
                ), color = "black", linewidth = 1.5) +
                geom_point(size = 5) +
                scale_color_gradient(low = "red", high = "blue") +
                theme_Publication_side() +
                ylab("Description") +
                xlab("Gene Ratio") +
                scale_x_continuous(expand = c(0, 0.01))
        } else {
            enrichment_plot <- ggplot() +
                theme_void()
        }
        ggsave(
            plot = enrichment_plot,
            file = here("QC", "results", "imgs", str_interp("SVA_incl_line${i}"), str_interp("ORA_PC${z}.png")),
            dpi=150, 
            width=10,
            height = 6

        )

    }

    pca_plot_df <- pca_data$rotated[, 1:20] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))
    for (k in 1:20) {
        pca_plot_df[, str_interp("PC${k}_explained_var")] <- pca_data$variance[k]
    }

    ##################
    # CALCULATE UMAP #
    ##################
    umap_data <- umap::umap(t(adjusted_counts))
    umap_plot_df <- umap_data$layout %>%
        data.frame() %>%
        magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
        rownames_to_column("sampleid") %>%
        left_join(., mdata)
    
    ##################
    # CALCULATE TSNE #
    ##################
    tsne_data <- Rtsne::Rtsne(t(adjusted_counts))
    tsne_plot_df <- tsne_data$Y %>%
            data.frame() %>%
            magrittr::set_colnames(c("TSNE1", "TSNE2")) %>%
            mutate("sampleid" = colnames(adjusted_counts)) %>%
            left_join(., mdata)


    ###########
    # HEATMAP #
    ###########
    association_output <- covar_correlation(
        df = pca_plot_df,
        covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
        components = paste0("PC", 1:15)
    )

    png(here("QC", "results", "imgs", str_interp("SVA_incl_line${i}"), "association_heatmap_corrected.png"),
        res = 150, width = 9, height = 5.5, units = "in"
    )
    association_output$heatmap
    dev.off()

    #############
    # PCA PLOTS #
    #############
    columns_to_plot <- c("state", "condition", "group", "ethnicity", "age", "batch1id")

    # create PC intervals
    PC_pairings <- lapply(1:floor(i/2), function(x) {
        c(2 * x - 1, 2 * x)
    })

    for (k in PC_pairings) {
        print(str_interp("Working on PC${k[1]} and PC${k[2]}"))
        plots <- lapply(columns_to_plot, function(x) {
            if (x == "group") {
                tmp_col = c("#fb867d", "#86f88f", "#7a8ff4", "#FA392A", "#0AB517", "#0b218d")
            } else {
                tmp_col = NA
            }
            
            expl_var1 <- round(pca_plot_df[1, str_interp("PC${k[1]}_explained_var")], 2)
            expl_var2 <- round(pca_plot_df[1, str_interp("PC${k[2]}_explained_var")], 2)
            p <- density_scatter(
                df = pca_plot_df,
                x_variable = str_interp("PC${k[1]}"),
                y_variable = str_interp("PC${k[2]}"),
                color_by = x,
                density_plot_ratio = 0.15,
                colors = tmp_col,
                pt_size = 1.2,
                xlab = str_interp("PC${k[1]}: ${expl_var1}% variance"),
                ylab = str_interp("PC${k[2]}: ${expl_var2}% variance")
            )
            p <- p +
                plot_annotation(
                    title = x,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
                )
            return(p)
        })
        final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
        ggsave(
            plot = final_plot,
            filename = here(
                "QC", "results", "imgs", str_interp("SVA_incl_line${i}"),
                str_interp("PCA_all_covariates_PC${k[1]}_PC${k[2]}_corrected.png")
            ),
            dpi = 150,
            width = 18,
            height = 11.5,
        )
    }

    ##############
    # UMAP PLOTS #
    ##############
    plots <- lapply(columns_to_plot, function(x) {
        if(x == "group"){
            tmp_col = c("#fb867d", "#86f88f", "#7a8ff4", "#FA392A", "#0AB517", "#0b218d")
        } else {
            tmp_col = NA
        }
        p <- density_scatter(
            df = umap_plot_df,
            x_variable = "UMAP1",
            y_variable = "UMAP2",
            colors = tmp_col,
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })

    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here("QC", "results", "imgs", str_interp("SVA_incl_line${i}"), "UMAP_all_covariates_corrected.png"),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
    
    ##############
    # TSNE PLOTS #
    ##############
    plots <- lapply(columns_to_plot, function(x) {
        p <- density_scatter(
            df = tsne_plot_df,
            x_variable = "TSNE1",
            y_variable = "TSNE2",
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })

    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here("QC", "results", "imgs", str_interp("SVA_incl_line${i}"), "TSNE_all_covariates_corrected.png"),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
    
}, mc.cores = 20) 
```


### Conclusion

**SVA**:
Results look actually very good. Not sure if I am overadjusting...
5 SVs look good.

```{r}
sessionInfo()
```





