---
title: "RNA seq batch correction"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 150
)
```


```{r}
show_batch <- function(dds, value, sampleids=NA, output_adjusted = TRUE, tsne=FALSE, col_name = "batch1id"){
    dds_subset <- dds[, colData(dds)[, col_name] == value]
    tmp_mdata <- mdata %>% data.frame() %>% filter(sampleid %in% dds_subset$sampleid)
    norm_counts <- assay(vst(dds_subset, blind = TRUE))
    mod <- model.matrix(~group, colData(dds_subset))
    mod0 <- model.matrix(~SV1 + SV2 + SV3 + SV4 + SV5, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/
    # adjust the counts using limma
    adjusted_counts <- limma::removeBatchEffect(norm_counts,
        batch = dds_subset$line,
        covariates = mod0,
        design = mod
    )

    pca_data_subset <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds_subset),
        removeVar = 0.4
    )

    pca_plot_df_subset <- pca_data_subset$rotated[, 1:10] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))

    for (i in 1:10) {
        pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
    }
   
    expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC1_explained_var")], 2)
    expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC2_explained_var")], 2)

    p_pca <- ggplot(pca_plot_df_subset, aes(x = PC1, y = PC2, color = group)) +
        geom_point(size = 3) +
        scale_color_manual(values = c("#ee794a", "#4cf73c", "#6071d3", "#FA392A", "#0AB517", "#0b218d")) +
        theme_Publication() +
        ggtitle(value) +
        xlab(str_interp("PC1: ${expl_var1}% variance")) +
        ylab(str_interp("PC2: ${expl_var2}% variance")) +
        guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
        ggrepel::geom_label_repel(
            data = pca_plot_df_subset %>% filter(sample_id %in% sampleids),
            aes(x = PC1, y = PC2, label = sample_id, size = 2.3),
            inherit.aes = FALSE, show.legend = FALSE,
            min.segment.length = 0,
            nudge_y = .2,
        )
    
    
    ########
    # UMAP #
    ########
    umap_data <- umap::umap(t(adjusted_counts))
    umap_plot_df <- umap_data$layout %>%
        data.frame() %>%
        magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))

    p_umap <- ggplot(umap_plot_df, aes(x = UMAP1, y = UMAP2, color = group)) +
        geom_point(size = 3) +
        scale_color_manual(values = c("#ee794a", "#4cf73c", "#6071d3", "#FA392A", "#0AB517", "#0b218d")) +
        theme_Publication() +
        ggtitle(value) +
        xlab("UMAP1") +
        ylab("UMAP2") +
        guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
        ggrepel::geom_label_repel(
            data = umap_plot_df %>% filter(sample_id %in% sampleids),
            aes(x = UMAP1, y = UMAP2, label = sample_id, size = 2.3),
            min.segment.length = 0,
            nudge_y = .2,
            inherit.aes = FALSE, show.legend = FALSE
        )
    
    if (tsne) {
        tsne_data <- Rtsne::Rtsne(t(adjusted_counts))
        tsne_plot_df <- tsne_data$Y %>%
            data.frame() %>%
            magrittr::set_colnames(c("TSNE1", "TSNE2")) %>%
            mutate("sample_id" = colnames(adjusted_counts)) %>%
            left_join(., mdata, by = c("sample_id" = "sampleid"))
        
        p_tsne <- ggplot(tsne_plot_df, aes(x = TSNE1, y = TSNE2, color = group)) +
            geom_point(size = 3) +
            scale_color_manual(values = c("#ee794a", "#4cf73c", "#6071d3", "#FA392A", "#0AB517", "#0b218d")) +
            theme_Publication() +
            ggtitle(value) +
            xlab("TSNE1") +
            ylab("TSNE2") +
            guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
            ggrepel::geom_label_repel(
                data = tsne_plot_df %>% filter(sample_id %in% sampleids),
                aes(x = TSNE1, y = TSNE2, label = sample_id, size = 2.3),
                min.segment.length = 0,
                nudge_y = .2,
                inherit.aes = FALSE, show.legend = FALSE
            )
        
    }
    

    if (!output_adjusted & tsne) {
        return(list(p_pca, p_umap, p_tsne))
    }

    if (!output_adjusted) {
        return(list(p_pca, p_umap))
    } 


    dds_updated <- dds_subset[, !dds_subset$sampleid %in% sampleids]
    print("Rerunning show_batch to remove highlighted sample")
    # i guess technically we would have to recalc sva, but this should be alright for the beginning
    updated_plots <- show_batch(dds = dds_updated, value = value, col_name = col_name, output_adjusted = F)
    final_plot <- wrap_plots(p_pca, p_umap, updated_plots[[1]], updated_plots[[2]], byrow=TRUE, ncol = 2)
    return(final_plot)
}

```

```{r, load libraries}
library(DESeq2)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(magrittr)
library(PCAtools)
library(umap)
library(patchwork)
library(sva)
library(clusterProfiler)
library(RUVSeq)
library(httpgd)
library(tidyverse)
library(Rtsne)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


## **Introduction**
If we look at our corrected UMAP, it seems like there are some mislabeled samples (M2 in a cloud of M1s).

Try and detect mislabeled samples for example samples that are actually unstimulated but labeled as stimulated. 
For this we will go through each biological group (i.e. M0_ctrl or M2_stim) and see if there are any outliers. 
For this, I will do an unadjusted PCA per library pool and then look if I see patterns. 

```{r}
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
mdata <- colData(dds) %>% data.frame()
sva_seq <- readRDS(here("QC", "results", "sva_incl_line_5SVs.rds"))
dds <- estimateSizeFactors(dds)
tmp_mdata <- cbind(mdata, sva_seq$sv %>% data.frame() %>% magrittr::set_colnames(paste0("SV", 1:5)))
norm_data <- assay(vst(dds, blind = TRUE))

mod <- model.matrix(~group, tmp_mdata)
formula <- as.formula(paste("~", paste0("SV", 1:5, collapse = " + ")))
mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

# adjust the counts using limma
adjusted_counts <- limma::removeBatchEffect(norm_data,
    batch = dds$line,
    covariates = mod0,
    design = mod
)

# anisogenic replicates
mdata <- mdata %>%
    mutate(group = paste(state, condition, sep="_")) 

cor_result <- lapply(unique(mdata$group), function(x){
    samples <- mdata %>%
        filter(group == x)  %>% 
        pull(sampleid)

    tmp_cnts <- adjusted_counts[, samples]
    cor_map <- cor(tmp_cnts, method = "spearman")
    return(cor_map)
})
names(cor_result) <- unique(mdata$group)

# some have very low correlation.
sapply(cor_result, min)
```


#### M0_ctrl
```{r}
col_fun = colorRamp2(c(0.5, 1), c("blue", "red"))
h <- Heatmap(cor_result$M0_ctrl,
    show_row_names = F,
    show_column_names = F,
    show_row_dend = F,
    show_column_dend = F,
    col = col_fun,
    row_km = 2,
    column_km = 2,
    name="R"
)

h <- draw(h)
# outliers_M0_ctrl <- column_order(draw(h))$`1`
# outliers_M0_ctrl <- rownames(cor_result$M0_ctrl)[outliers_M0_ctrl]
apply(cor_result$M0_ctrl, 2, mean) %>% sort(decreasing = T)

# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M0_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M0_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)

# MM_oxLDL9752143 is very different from the rest... remove it
M0_ctrl_outlier <- c("MM_oxLDL9752143")

```

#### M1_ctrl
```{r}
col_fun = colorRamp2(c(0.7, 1), c("blue", "red"))
h <- Heatmap(cor_result$M1_ctrl,
    show_row_names = F,
    show_column_names = F,
    show_row_dend = F,
    show_column_dend = F,
    col = col_fun,
    row_km = 2,
    column_km = 2,
    name="R"
)
draw(h)

# looks like there is really just one sample which is very different
outliers_M1_ctrl <- apply(cor_result$M1_ctrl, 2, mean) %>% which.min() %>% names()

# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M1_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M1_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)

# MM_oxLDL9752115 is very different from the rest, prob. because of very low libsize 
M1_ctrl_outlier <- c("MM_oxLDL9752115")

```

#### M2_ctrl
```{r}
col_fun = colorRamp2(c(0.7, 1), c("blue", "red"))
h <- Heatmap(cor_result$M2_ctrl,
    show_row_names = F,
    show_column_names = F,
    show_row_dend = F,
    show_column_dend = F,
    col = col_fun,
    row_km = 2,
    column_km = 2,
    name="R"
)
draw(h)

# looks like there is really just one sample which is very different
outliers_M2_ctrl <- apply(cor_result$M2_ctrl, 2, mean) %>% which.min() %>% names()

# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M2_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M2_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
# low libsize again. 
M2_ctrl_outlier <- c("MM_oxLDL9752225")

```

#### M0_stim
```{r}
col_fun = colorRamp2(c(0.7, 1), c("blue", "red"))
h <- Heatmap(cor_result$M0_stim,
    show_row_names = F,
    show_column_names = F,
    show_row_dend = F,
    show_column_dend = F,
    col = col_fun,
    row_km = 2,
    column_km = 2,
    name="R"
)
draw(h)

# looks like there is really just one sample which is very different
outliers_M0_stimm <- apply(cor_result$M0_stim, 2, mean) %>% which.min() %>% names()



# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M0_stim)],
    metadata = colData(dds)[rownames(cor_result$M0_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)

# MM_oxLDL7992361
mdata %>% filter(sampleid=="MM_oxLDL7992361")
M0_oxldl_outlier <- c("MM_oxLDL7992361")
```

#### M1_stim
```{r}
col_fun = colorRamp2(c(0.7, 1), c("blue", "red"))
h <- Heatmap(cor_result$M1_stim,
    show_row_names = F,
    show_column_names = F,
    show_row_dend = F,
    show_column_dend = F,
    col = col_fun,
    row_km = 2,
    column_km = 2,
    name="R"
)
draw(h)

# looks like there is really just one sample which is very different
outliers_M1_stimm <- apply(cor_result$M1_stim, 2, mean) %>% which.min() %>% names()


# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M1_stim)],
    metadata = colData(dds)[rownames(cor_result$M1_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)

M1_oxldl_outlier <- "MM_oxLDL7734278"
```

#### M2_stim
```{r}
col_fun = colorRamp2(c(0.7, 1), c("blue", "red"))
h <- Heatmap(cor_result$M2_stim,
    show_row_names = F,
    show_column_names = F,
    show_row_dend = F,
    show_column_dend = F,
    col = col_fun,
    row_km = 2,
    column_km = 2,
    name="R"
)
draw(h)

# This looks just fine. No outliers
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M2_stim)],
    metadata = colData(dds)[rownames(cor_result$M2_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)


```

### Bringing it all together
```{r}
outliers <- c(
    M0_ctrl_outlier, M1_ctrl_outlier, M2_ctrl_outlier,
    M0_oxldl_outlier, M1_oxldl_outlier 
)

dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
dds_filtered <- dds[, !dds$sampleid  %in% outliers]
saveRDS(dds_filtered, here("QC", "results", "dds_sva_incl_line_libsize_cleaned.rds"))
```

### Recalculate SVA
This actually did not change much... Still some samples that seem to be in the "wrong" group.

```{r calculate }
# load in DESeq2dataset where the libsize outliers are already removed.
# for creation see script **01_initial_QC.rmd**
dds <- readRDS(here("QC", "results", "dds_sva_incl_line_libsize_cleaned.rds"))
dds <- estimateSizeFactors(dds)
mdata <- colData(dds) %>% data.frame()
norm_counts <- counts(dds, norm=TRUE)
vsd_data <- vst(dds, blind=T)
mod <- model.matrix(~line+group, colData(dds))
mod0 <- model.matrix(~line, colData(dds))

# runs long, even with parallization
# dummy_ <- parallel::mclapply(3:10, function(i) {
#     svseq <- svaseq(norm_counts, mod, mod0, n.sv = i)
#     saveRDS(svseq, here("QC", "results", str_interp("sva_incl_line_libsize_cleaned_${i}SVs.rds")))
# }, mc.cores=20)

hs_msigdb_df <- msigdbr::msigdbr(species = "Homo sapiens")

gene_sets <- hs_msigdb_df %>%
    filter(gs_subcat == "GO:BP" | gs_cat == "H" | gs_subcat == "CP:KEGG") %>%
    select(gs_name, gene_symbol)

dummy_ <- parallel::mclapply(3:10, function(i){
    dir.create(here("QC", "results", "imgs", str_interp("SVA_incl_line_libsize_cleaned${i}")), showWarnings = FALSE)
    svseq <- readRDS(here("QC", "results", str_interp("sva_incl_line_libsize_cleaned_${i}SVs.rds")))
    SVs <- svseq$sv
    colnames(SVs) <- paste0("SV", 1:i)
    tmp_mdata <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)

    # use limmas removeBatchEffects to emulate what the linear model will eventually do.
    norm_data <- assay(vsd_data)
    mod <- model.matrix(~group, tmp_mdata)
    # Create the formula dynamically
    formula <- as.formula(paste("~", paste0("SV", 1:i, collapse = " + ")))
    mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

    # adjust the counts using limma
    adjusted_counts <- limma::removeBatchEffect(norm_data,
        batch = vsd_data$line,
        covariates = mod0,
        design = mod
    )
    # recalculate PCA with adjusted counts
    pca_data <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds),
        removeVar = 0.4
    )

    # Extract the genes which most influence PC1 and PC2 and see if they are enriched in any biological domain:
    for (z in 1:2) {
        print(paste("Running ORA for PC", z))
        # extract top 300 genes which most influence PC
        top_loadings_gene_names <- pca_data$xvars[order(abs(pca_data$loadings[, paste0("PC", z)]), decreasing = T)[1:300]]

        enrichment_result <- enricher(
            gene = top_loadings_gene_names,
            universe = rownames(dds),
            minGSSize = 5,
            maxGSSize = 300,
            TERM2GENE = gene_sets
        )

        # for each of the biological domains, grep the top 3 sig. terms
        # for each of the biological domains, grep the top 3 sig. terms
        sig_results <- enrichment_result@result %>%
            mutate(domain = str_extract(Description, "^[^_]*")) %>%
            mutate(ratio_numeric = sapply(.$GeneRatio, function(x) {
                eval(
                    str2expression(x)
                )
            })) %>%
            filter(p.adjust < 0.05) %>%
            group_by(domain) %>%
            slice_min(., order_by = p.adjust, n = 3) %>%
            mutate(Description_short = shorten_strings(Description, 30))
        
        if (nrow(sig_results) > 0) {
            enrichment_plot <- ggplot(sig_results, aes(x = ratio_numeric, y = Description_short, color = p.adjust)) +
                geom_segment(aes(
                    x = 0, xend = ratio_numeric, y = Description_short, yend = Description_short
                ), color = "black", linewidth = 1.5) +
                geom_point(size = 5) +
                scale_color_gradient(low = "red", high = "blue") +
                theme_Publication_side() +
                ylab("Description") +
                xlab("Gene Ratio") +
                scale_x_continuous(expand = c(0, 0.01))
        } else {
            enrichment_plot <- ggplot() +
                theme_void()
        }
        ggsave(
            plot = enrichment_plot,
            file = here(
                "QC", "results", "imgs", str_interp("SVA_incl_line_libsize_cleaned${i}"),
                str_interp("ORA_PC${z}.png")
            ),
            dpi = 150,
            width = 10,
            height = 6
        )
    }


    pca_plot_df <- pca_data$rotated[, 1:20] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))
    for (k in 1:20) {
        pca_plot_df[, str_interp("PC${k}_explained_var")] <- pca_data$variance[k]
    }

    ##################
    # CALCULATE UMAP #
    ##################
    umap_data <- umap::umap(t(adjusted_counts))
    umap_plot_df <- umap_data$layout %>%
        data.frame() %>%
        magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
        rownames_to_column("sampleid") %>%
        left_join(., mdata)
    
    ##################
    # CALCULATE TSNE #
    ##################
    tsne_data <- Rtsne::Rtsne(t(adjusted_counts))
    tsne_plot_df <- tsne_data$Y %>%
            data.frame() %>%
            magrittr::set_colnames(c("TSNE1", "TSNE2")) %>%
            mutate("sampleid" = colnames(adjusted_counts)) %>%
            left_join(., mdata)


    ###########
    # HEATMAP #
    ###########
    association_output <- covar_correlation(
        df = pca_plot_df,
        covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
        components = paste0("PC", 1:15)
    )

    png(here("QC", "results", "imgs", str_interp("SVA_incl_line_libsize_cleaned${i}"), "association_heatmap_corrected.png"),
        res = 150, width = 9, height = 5.5, units = "in"
    )
    association_output$heatmap
    dev.off()

    #############
    # PCA PLOTS #
    #############
    columns_to_plot <- c("state", "condition", "group", "ethnicity", "age", "batch1id")

    # create PC intervals
    PC_pairings <- lapply(1:floor(i/2), function(x) {
        c(2 * x - 1, 2 * x)
    })

    for (k in PC_pairings) {
        print(str_interp("Working on PC${k[1]} and PC${k[2]}"))
        plots <- lapply(columns_to_plot, function(x) {
            expl_var1 <- round(pca_plot_df[1, str_interp("PC${k[1]}_explained_var")], 2)
            expl_var2 <- round(pca_plot_df[1, str_interp("PC${k[2]}_explained_var")], 2)
            p <- density_scatter(
                df = pca_plot_df,
                x_variable = str_interp("PC${k[1]}"),
                y_variable = str_interp("PC${k[2]}"),
                color_by = x,
                density_plot_ratio = 0.15,
                pt_size = 1.2,
                xlab = str_interp("PC${k[1]}: ${expl_var1}% variance"),
                ylab = str_interp("PC${k[2]}: ${expl_var2}% variance")
            )
            p <- p +
                plot_annotation(
                    title = x,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
                )
            return(p)
        })
        final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
        ggsave(
            plot = final_plot,
            filename = here(
                "QC", "results", "imgs", str_interp("SVA_incl_line_libsize_cleaned${i}"),
                str_interp("PCA_all_covariates_PC${k[1]}_PC${k[2]}_corrected.png")
            ),
            dpi = 150,
            width = 18,
            height = 11.5,
        )
    }

    ##############
    # UMAP PLOTS #
    ##############
    plots <- lapply(columns_to_plot, function(x) {
        p <- density_scatter(
            df = umap_plot_df,
            x_variable = "UMAP1",
            y_variable = "UMAP2",
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })

    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here("QC", "results", "imgs", str_interp("SVA_incl_line_libsize_cleaned${i}"), "UMAP_all_covariates_corrected.png"),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
    
    ##############
    # TSNE PLOTS #
    ##############
    plots <- lapply(columns_to_plot, function(x) {
        p <- density_scatter(
            df = tsne_plot_df,
            x_variable = "TSNE1",
            y_variable = "TSNE2",
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })

    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here("QC", "results", "imgs", str_interp("SVA_incl_line_libsize_cleaned${i}"), "TSNE_all_covariates_corrected.png"),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
    
}, mc.cores = 20) 



```

### Create PC and UMAP plot per batch
Some look very nice, others very messy... 

```{r calculate SVs}
# load in DESeq2dataset where the libsize outliers are already removed.
dds <- readRDS(here("QC", "results", "dds_sva_incl_line_libsize_cleaned.rds"))
dds <- estimateSizeFactors(dds)
mdata <- colData(dds) %>% data.frame()
sva_seq <- readRDS(here("QC", "results", "sva_incl_line_libsize_cleaned_5SVs.rds"))
mdata <- cbind(mdata, sva_seq$sv %>% data.frame() %>% magrittr::set_colnames(paste0("SV", 1:5)))


batchwise_plots <- parallel::mclapply(droplevels(sort(unique(dds$batch1id))), function(batch) {
    dds_subset <- dds[, dds$batch1id == batch]
    if(ncol(dds_subset)<15){
        return()
    }
    tmp_mdata <- mdata %>% filter(sampleid %in% dds_subset$sampleid)
    norm_counts <- assay(vst(dds_subset, blind = TRUE))
    mod <- model.matrix(~group, colData(dds_subset))
    formula <- as.formula(paste("~", paste0("SV", 1:5, collapse = " + ")))
    mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/
    # adjust the counts using limma
    adjusted_counts <- limma::removeBatchEffect(norm_counts,
        batch = dds_subset$line,
        covariates = mod0,
        design = mod
    )

    pca_data_subset <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds_subset),
        removeVar = 0.4
    )
    pca_plot_df_subset <- pca_data_subset$rotated[, 1:10] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))

    for (i in 1:10) {
        pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
    }

    expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC1_explained_var")], 2)
    expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC2_explained_var")], 2)
    p <- density_scatter(
        df = pca_plot_df_subset,
        x_variable = "PC1",
        y_variable = "PC2",
        color_by = "group",
        color = c("#ee794a", "#4cf73c", "#6071d3", "#FA392A", "#0AB517", "#0b218d"),
        density_plot_ratio = 0,
        pt_size = 2,
        xlab = str_interp("PC1: ${expl_var1}% variance"),
        ylab = str_interp("PC2: ${expl_var2}% variance")
    )

    p_pca <- p +
        plot_annotation(
            title = batch,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )

    ########
    # UMAP #
    ########
    umap_data <- umap::umap(t(adjusted_counts))
    umap_plot_df <- umap_data$layout %>%
        data.frame() %>%
        magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
        rownames_to_column("sampleid") %>%
        left_join(., mdata)


    p <- density_scatter(
        df = umap_plot_df,
        x_variable = str_interp("UMAP1"),
        y_variable = str_interp("UMAP2"),
        color_by = "group",
        color = c("#ee794a", "#4cf73c", "#6071d3", "#FA392A", "#0AB517", "#0b218d"),
        density_plot_ratio = 0,
        pt_size = 2
    )
    p_umap <- p +
        plot_annotation(
            title = batch,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )

    return(list("pca" = p_pca, "umap" = p_umap))
}, mc.cores=20)


batchwise_plots <- batchwise_plots[lengths(batchwise_plots) != 0]
pcas <- sapply(batchwise_plots, function(x){x[[1]]})
umaps <- sapply(batchwise_plots, function(x){x[[2]]})


pcas_final <- cowplot::plot_grid(plotlist = pcas, ncol=5)
umaps_final <- cowplot::plot_grid(plotlist = umaps, ncol=5)

ggsave(
    plot = pcas_final,
    filename = here("QC", "results", "imgs",  "SVA_incl_line_libsize_cleaned5", "batchwise_PCA_combined.png"),
    dpi = 450,
    width = 24,
    height = 15,
    scale=1.15
)

ggsave(
    plot = umaps_final,
    filename = here("QC", "results", "imgs",  "SVA_incl_line_libsize_cleaned5", "batchwise_UMAP_combined.png"),
    dpi = 450,
    width = 24,
    height = 15,
    scale=1.15
)

```


### Go through all the batches and investigate outliers
```{r}
dds <- readRDS(here("QC", "results", "dds_sva_incl_line_libsize_cleaned.rds"))
dds <- estimateSizeFactors(dds)
mdata <- colData(dds) %>% data.frame()
sva_seq <- readRDS(here("QC", "results", "sva_incl_line_libsize_cleaned_5SVs.rds"))
mdata <- cbind(mdata, sva_seq$sv %>% data.frame() %>% magrittr::set_colnames(paste0("SV", 1:5)))
```

#### 65609
```{r}
# 1 M1_ctrl sample clusters to the M0_ctrl.
# visible in both PCA and UMAP
# sample is MM_oxLDL7734271

# belongs to the "corn1" line, which are all in this batch... so I think that it is unlikely that there is an actual swap
# maybe it just behaves very different or the differentiation to M1 did not work out. Library size is also normal
mdata %>% filter(sampleid=="MM_oxLDL7734271")
swaps_65609 <- c("MM_oxLDL7734271")

show_batch(
    dds = dds,
    value = "65609",
    output_adjusted = F,
    sampleids = swaps_65609,
    tsne = TRUE
) %>% wrap_plots()


show_batch(
    dds = dds,
    value = "65609",
    sampleids = swaps_65609,
    output_adjusted = T
)

show_batch(
    dds = dds,
    value = "65609",
    sampleids = swaps_65609,
    output_adjusted = T
) 
mdata %>% filter(batch1id == "65609") %>% pull(oxldlset) %>% droplevels() %>%  table()
show_batch(
    dds = dds,
    value = "13",
    col_name = "oxldlset",
    sampleids = swaps_65609,
    output_adjusted = T
) 
```

#### 69511
```{r}
# here it looks like the oxldl treatment was weak..
# cluster according to celltype a lot
show_batch(
    dds = dds,
    value = "69511",
    output_adjusted = F
) 
mdata %>% filter(batch1id == "69511") %>% pull(oxldlset) %>% droplevels() %>%  table()

# we observe the same for cluster 19 actually.. UMAP looks fine though. 
# Also one M0_ctrl looks like an outlier (MM_oxLDL7954434)
show_batch(
    dds = dds,
    value = "19",
    col_name = "oxldlset",
    output_adjusted = F
)  %>% wrap_plots()


show_batch(
    dds = dds,
    value = "19",
    col_name = "oxldlset",
    sampleids = "MM_oxLDL7954434",
    output_adjusted = T
)

# for oxldl set 23 it looks good!
show_batch(
    dds = dds,
    value = "23",
    col_name = "oxldlset",
    output_adjusted = F
)  %>% wrap_plots()



swaps_69511 <- c("MM_oxLDL7954434")
# still one M2_stim outlier: MM_oxLDL7954415
show_batch(
    dds = dds,
    value = "69511",
    sampleids = swaps_69511,
    output_adjusted = T
)
swaps_69511 <- c("MM_oxLDL7954434", "MM_oxLDL7954415")

show_batch(
    dds = dds,
    value = "69511",
    sampleids = swaps_69511,
    output_adjusted = T
)
```


#### 70163
```{r}
# one M0_ctrl clusters far away from the other samples (MM_oxLDL7992360). Might be because of small sample size (135k reads)
# in the umap it actually looks fine

# however MM_oxLDL7992342 looks quite different
show_batch(
    dds = dds,
    value = "70163",
    sampleids = c("MM_oxLDL7992360", "MM_oxLDL7992342"),
    output_adjusted = T
)

# honestly for 70163 it does not look bad, we might remove to much here..
swaps_70163 <- c("MM_oxLDL7992360", "MM_oxLDL7992342")
```


#### 70601
```{r}
# on the umap looks like there are 2 outlier.
# both are M0_ctrl -> MM_oxLDL8032539, MM_oxLDL8032557
show_batch(
    dds = dds,
    value = "70601",
    output_adjusted = F
)

# looks fine!
```

#### 71670
```{r}
# on the umap looks like there are 2 outlier.
# both are M0_ctrl -> MM_oxLDL8032539, MM_oxLDL8032557
show_batch(
    dds = dds,
    value = "71670",
    output_adjusted = F
) 

mdata %>%
    filter(batch1id == 71670) %>%
    pull(oxldlset) %>%
    droplevels() %>%
    table()


# always one is misplaced, might be a true sample swap.
swaps_71670 <- c(
    "MM_oxLDL8121203", "MM_oxLDL8121201", "MM_oxLDL8121200",
    "MM_oxLDL8121204", "MM_oxLDL8121205", "MM_oxLDL8121202"
)

show_batch(
    dds = dds,
    value = "71670",
    sampleids = swaps_71670,
    output_adjusted = F,
    tsne=TRUE
) %>% wrap_plots()
# ALL of these are the same cell line. It is highly likely that they actually have the wrong annotation.
mdata %>% filter(sampleid %in% swaps_71670) %>% pull(line)

# in this case it makes probably more sense to do a reassignment rather than just removing them 
```

#### 79551
```{r}
# 2 extreme outliers for PCA, but UMAP looks fine: MM_oxLDL9752083 and MM_oxLDL9752107
show_batch(
    dds = dds,
    value = "79551",
    output_adjusted = F,
    tsne = TRUE
) %>% wrap_plots()

show_batch(
    dds = dds,
    value = "79551",
    output_adjusted = TRUE,
    sampleids = c("MM_oxLDL9752083", "MM_oxLDL9752107")
)

mdata %>% filter(sampleid %in% c("MM_oxLDL9752083", "MM_oxLDL9752107"))

swaps_79551 <- c("MM_oxLDL9752083", "MM_oxLDL9752107")


```


#### 79552
```{r}
# there might be one outlier M2_ctrl sample, but it is very hard to say.. only in UMAP  and tsne visible
# outlier is MM_oxLDL9752321
show_batch(
    dds = dds,
    value = "79552",
    output_adjusted = F,
    sampleids = c("MM_oxLDL9752321"),
    tsne = TRUE
) %>% wrap_plots()


# all individual oxldlsets look great
mdata %>%
    filter(batch1id == 79552) %>%
    pull(oxldlset) %>%
    droplevels() %>%
    table()

swaps_79552 <- c("MM_oxLDL9752321")
show_batch(
    dds = dds,
    value = "79552",
    output_adjusted = T,
    sampleids = swaps_79552,
) %>% wrap_plots()

```


#### 79583
```{r}
show_batch(
    dds = dds,
    value = "79583",
    output_adjusted = F,
    tsne = FALSE
) %>% wrap_plots()

swaps_79552 <- c()
# looks good!
```




#### 79704
```{r}
# this one is actually very complicated... Looks very messy
show_batch(
    dds = dds,
    value = "79704",
    sampleids = NA,
    output_adjusted = F
)
# look at the oxldl sets that are processed in it. One thing that stands out that this is composed of 
# 5 different oxldlsets, which is very unusual.. usually just 2. 
mdata %>% filter(batch1id == "79704") %>% pull(oxldlset) %>% droplevels() %>% table()
# 30 looks a bit off... M2 ctrl close to M1 stimulated. sampleID: MM_oxLDL9756613
# also MM_oxLDL9756614 (M2_stim) is close to M0 ctrl cloud
show_batch(
    dds = dds,
    value = "30",
    col_name = "oxldlset",
    sampleids = c("MM_oxLDL9756613", "MM_oxLDL9756614"),
    output_adjusted = T
)


show_batch(
    dds = dds,
    value = "30",
    col_name = "oxldlset",
    sampleids = mdata %>% filter(oxldlset == 30) %>% pull(sampleid),
    output_adjusted = F
)

# 31 looks very good
# 33 looks a bit off... the stimulated ones are quite close together. Also M2 ctrl clusters to a M0_ctrl. sample is MM_oxLDL9752207
show_batch(
    dds = dds,
    value = "33",
    col_name = "oxldlset",
    sampleids = "MM_oxLDL9752207",
    output_adjusted = TRUE
)
## 34 looks decent
swaps_79704 <- c("MM_oxLDL9756613", "MM_oxLDL9756614", "MM_oxLDL9752207")

# this still looks pretty bad...
show_batch(
    dds = dds,
    value = "79704",
    sampleids = swaps_79704,
    output_adjusted = T
)
```


```{r}
sessionInfo()
```





