---
title: "RNA seq batch correction"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 150
)
```

```{r, load libraries}
library(DESeq2)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(magrittr)
library(PCAtools)
library(umap)
library(patchwork)
library(sva)
library(clusterProfiler)
library(RUVSeq)
library(httpgd)
library(tidyverse)
library(Rtsne)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


## **Introduction**
In 02b_SVA_with_known_covar_outlier_detection.rmd we have looked at each batch and have removed (or corrected) wrong samples. 
Here we will rerun SVA one final time and replot the UMAPs. I expect the results will not change much, just a bit cleaner

```{r}
dds <- readRDS(here("QC", "results", "dds_final.rds"))
mdata <- colData(dds) %>% data.frame()
norm_data <- assay(vst(dds, blind = TRUE))

mod <- model.matrix(~ line + group, colData(dds))
mod0 <- model.matrix(~line, colData(dds))

# # runs long, even with parallization
# svseq <- svaseq(norm_data, mod, mod0, n.sv = 5)
# saveRDS(svseq, here("QC", "results", "sva_5_final.rds"))

svseq <- readRDS(here("QC", "results", "sva_5_final.rds"))
SVs <- svseq$sv
colnames(SVs) <- paste0("SV", 1:5)
mdata <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)


# use limmas removeBatchEffects to emulate what the linear model will eventually do.
mod <- model.matrix(~group, mdata)
# Create the formula dynamically
formula <- as.formula(paste("~", paste0("SV", 1:5, collapse = " + ")))
mod0 <- model.matrix(formula, mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

# adjust the counts using limma
adjusted_counts <- limma::removeBatchEffect(norm_data,
    batch = dds$line,
    covariates = mod0,
    design = mod
)
cor_result <- lapply(unique(mdata$group), function(x){
    samples <- mdata %>%
        filter(group == x)  %>% 
        pull(sampleid)

    tmp_cnts <- adjusted_counts[, samples]
    cor_map <- cor(tmp_cnts, method = "spearman")
    return(cor_map)
})
names(cor_result) <- unique(mdata$group)

sapply(cor_result, min)
```


#### M0_ctrl
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M0_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M0_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset,
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "batch1id",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
```

#### M1_ctrl
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M1_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M1_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset ,
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "batch1id",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
```

#### M2_ctrl
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M2_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M2_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset,
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "batch1id",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
```

#### M0_stim
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M0_stim)],
    metadata = colData(dds)[rownames(cor_result$M0_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset,
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "batch1id",
    density_plot_ratio = 0,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
```

#### M1_stim
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M1_stim)],
    metadata = colData(dds)[rownames(cor_result$M1_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset ,
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "batch1id",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
```

#### M2_stim
```{r}
# This looks just fine. No outliers
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M2_stim)],
    metadata = colData(dds)[rownames(cor_result$M2_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
density_scatter(
    df = pca_plot_df_subset ,
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "batch1id",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
```

### Replot the final data
```{r recreate the plots one final time}
hs_msigdb_df <- msigdbr::msigdbr(species = "Homo sapiens")

gene_sets <- hs_msigdb_df %>%
    filter(gs_subcat == "GO:BP" | gs_cat == "H" | gs_subcat == "CP:KEGG") %>%
    select(gs_name, gene_symbol)


# recalculate PCA with adjusted counts
pca_data <- PCAtools::pca(adjusted_counts,
    metadata = colData(dds),
    removeVar = 0.4
)

# Extract the genes which most influence PC1 and PC2 and see if they are enriched in any biological domain:
for (z in 1:2) {
    print(paste("Running ORA for PC", z))
    # extract top 300 genes which most influence PC
    top_loadings_gene_names <- pca_data$xvars[order(abs(pca_data$loadings[, paste0("PC", z)]), decreasing = T)[1:300]]

    enrichment_result <- enricher(
        gene = top_loadings_gene_names,
        universe = rownames(dds),
        minGSSize = 5,
        maxGSSize = 300,
        TERM2GENE = gene_sets
    )

    # for each of the biological domains, grep the top 3 sig. terms
    # for each of the biological domains, grep the top 3 sig. terms
    sig_results <- enrichment_result@result %>%
        mutate(domain = str_extract(Description, "^[^_]*")) %>%
        mutate(ratio_numeric = sapply(.$GeneRatio, function(x) {
            eval(
                str2expression(x)
            )
        })) %>%
        filter(p.adjust < 0.05) %>%
        group_by(domain) %>%
        slice_min(., order_by = p.adjust, n = 3) %>%
        mutate(Description_short = shorten_strings(Description, 30))
    
    if (nrow(sig_results) > 0) {
        enrichment_plot <- ggplot(sig_results, aes(x = ratio_numeric, y = Description_short, color = p.adjust)) +
            geom_segment(aes(
                x = 0, xend = ratio_numeric, y = Description_short, yend = Description_short
            ), color = "black", linewidth = 1.5) +
            geom_point(size = 5) +
            scale_color_gradient(low = "red", high = "blue") +
            theme_Publication_side() +
            ylab("Description") +
            xlab("Gene Ratio") +
            scale_x_continuous(expand = c(0, 0.01))
    } else {
        enrichment_plot <- ggplot() +
            theme_void()
    }
    ggsave(
        plot = enrichment_plot,
        file = here(
            "QC", "results", "imgs", str_interp("SVA5_final"),
            str_interp("ORA_PC${z}.png")
        ),
        dpi = 150,
        width = 10,
        height = 6
    )
}


pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (k in 1:20) {
    pca_plot_df[, str_interp("PC${k}_explained_var")] <- pca_data$variance[k]
}

##################
# CALCULATE UMAP #
##################
umap_data <- umap::umap(t(adjusted_counts))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>%
    left_join(., mdata)

##################
# CALCULATE TSNE #
##################
tsne_data <- Rtsne::Rtsne(t(adjusted_counts))
tsne_plot_df <- tsne_data$Y %>%
        data.frame() %>%
        magrittr::set_colnames(c("TSNE1", "TSNE2")) %>%
        mutate("sampleid" = colnames(adjusted_counts)) %>%
        left_join(., mdata)


###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
    components = paste0("PC", 1:15)
)

png(here("QC", "results", "imgs", str_interp("SVA5_final"), "association_heatmap_corrected.png"),
    res = 150, width = 9, height = 5.5, units = "in"
)
association_output$heatmap
dev.off()

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "condition", "group", "ethnicity", "age", "batch1id")

# create PC intervals
PC_pairings <- lapply(1:floor(5/2), function(x) {
    c(2 * x - 1, 2 * x)
})

for (k in PC_pairings) {
    print(str_interp("Working on PC${k[1]} and PC${k[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${k[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${k[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${k[1]}"),
            y_variable = str_interp("PC${k[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${k[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${k[2]}: ${expl_var2}% variance")
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "QC", "results", "imgs", str_interp("SVA5_final"),
            str_interp("PCA_all_covariates_PC${k[1]}_PC${k[2]}_corrected.png")
        ),
        dpi = 150,
        width = 18,
        height = 11.5,
    )
}

##############
# UMAP PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", str_interp("SVA5_final"), "UMAP_all_covariates_corrected.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)

##############
# TSNE PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = tsne_plot_df,
        x_variable = "TSNE1",
        y_variable = "TSNE2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", str_interp("SVA5_final"), "TSNE_all_covariates_corrected.png"),
    dpi = 150,
    width = 18,
    height = 11.5,
)

```

### Create PC and UMAP plot per batch
Some look very nice, others very messy... 

```{r calculate SVs}
batchwise_plots <- parallel::mclapply(droplevels(sort(unique(dds$batch1id))), function(batch) {
    dds_subset <- dds[, dds$batch1id == batch]
    if(ncol(dds_subset)<15){
        return()
    }
    tmp_mdata <- mdata %>% filter(sampleid %in% dds_subset$sampleid)
    norm_counts <- assay(vst(dds_subset, blind = TRUE))
    mod <- model.matrix(~group, colData(dds_subset))
    formula <- as.formula(paste("~", paste0("SV", 1:5, collapse = " + ")))
    mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/
    # adjust the counts using limma
    adjusted_counts <- limma::removeBatchEffect(norm_counts,
        batch = dds_subset$line,
        covariates = mod0,
        design = mod
    )

    pca_data_subset <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds_subset),
        removeVar = 0.4
    )
    pca_plot_df_subset <- pca_data_subset$rotated[, 1:10] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))

    for (i in 1:10) {
        pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
    }

    expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC1_explained_var")], 2)
    expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC2_explained_var")], 2)
    p <- density_scatter(
        df = pca_plot_df_subset,
        x_variable = "PC1",
        y_variable = "PC2",
        color_by = "group",
        color = c("#ee794a", "#4cf73c", "#6071d3", "#FA392A", "#0AB517", "#0b218d"),
        density_plot_ratio = 0,
        pt_size = 2,
        xlab = str_interp("PC1: ${expl_var1}% variance"),
        ylab = str_interp("PC2: ${expl_var2}% variance")
    )

    p_pca <- p +
        plot_annotation(
            title = batch,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )

    ########
    # UMAP #
    ########
    umap_data <- umap::umap(t(adjusted_counts))
    umap_plot_df <- umap_data$layout %>%
        data.frame() %>%
        magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
        rownames_to_column("sampleid") %>%
        left_join(., mdata)


    p <- density_scatter(
        df = umap_plot_df,
        x_variable = str_interp("UMAP1"),
        y_variable = str_interp("UMAP2"),
        color_by = "group",
        color = c("#ee794a", "#4cf73c", "#6071d3", "#FA392A", "#0AB517", "#0b218d"),
        density_plot_ratio = 0,
        pt_size = 2
    )
    p_umap <- p +
        plot_annotation(
            title = batch,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )

    return(list("pca" = p_pca, "umap" = p_umap))
}, mc.cores=20)


batchwise_plots <- batchwise_plots[lengths(batchwise_plots) != 0]
pcas <- sapply(batchwise_plots, function(x){x[[1]]})
umaps <- sapply(batchwise_plots, function(x){x[[2]]})


pcas_final <- cowplot::plot_grid(plotlist = pcas, ncol=5)
umaps_final <- cowplot::plot_grid(plotlist = umaps, ncol=5)

ggsave(
    plot = pcas_final,
    filename = here("QC", "results", "imgs",  "SVA5_final", "batchwise_PCA_combined.png"),
    dpi = 450,
    width = 24,
    height = 15,
    scale=1.15
)

ggsave(
    plot = umaps_final,
    filename = here("QC", "results", "imgs",  "SVA5_final", "batchwise_UMAP_combined.png"),
    dpi = 450,
    width = 24,
    height = 15,
    scale=1.15
)

```


```{r}
sessionInfo()
```





