---
title: "RNA seq batch correction using metadata"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 150
)
```

## **Introduction**
After doing some initial QC we detect strong batch effects try to correct them using information from the meta data.
Here we correct it using the metadata and filter out samples with less than 5M reads.

```{r, load libraries}
library(DESeq2)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(magrittr)
library(PCAtools)
library(umap)
library(patchwork)
library(sva)
library(clusterProfiler)
library(RUVSeq)
library(httpgd)
library(tidyverse)
library(Rtsne)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```

```{r}
calc_pairwise_cor <- function(mat, outliers, condition_){
    keep_samples <- mdata %>% filter(group == condition_)  %>% pull(sampleid)
    tmp_cnts <- mat[, keep_samples]
    tmp_cnts <- tmp_cnts[, !colnames(tmp_cnts) %in% outliers]
    pairwise_correlation <- cor(tmp_cnts, method = "spearman")
    h <- Heatmap(pairwise_correlation,
        show_row_names = F,
        show_column_names = F,
        show_row_dend = F,
        show_column_dend = F,
        col = col_fun,
        name="R"
    ) %>%
        ggplotify::as.ggplot()
    return(list(
        heatmap = h,
        correlation_matrix = pairwise_correlation
    ))
}
```


## Adjust count data
```{r}
# for creation see script **01_initial_QC.rmd**
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
dds_filtered <- dds[, dds$libsize > 5e6]

mdata <- colData(dds_filtered) %>% data.frame()

# regress out donor effect (cell line)
# use limmas removeBatchEffects to emulate what the linear model will eventually do.
mod <- model.matrix(~group, colData(dds_filtered))
norm_cnts <- assay(vst(dds_filtered))
adjusted_counts <- limma::removeBatchEffect(norm_cnts,
    batch = mdata$line, 
    design = mod
)

#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(adjusted_counts,
    metadata = colData(dds_filtered),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., colData(dds_filtered) %>% data.frame(), by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

PC_pairings <- lapply(1:3, function(x){
    c(2*x-1,2*x)
})
pca_plots <- lapply(PC_pairings, function(i) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(c("state", "condition", "sex", "ethnicity", "age", "batch1id"), function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
})

# honestly I barely notice any difference... W
pca_plots[[1]]
pca_plots[[2]]


##################
# CALCULATE UMAP #
##################
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
umap_data <- umap(t(adjusted_counts))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>%
    left_join(., mdata)

umap_plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = umap_plots, ncol = 3)
final_plot

##################
# CALCULATE TSNE #
##################
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
tsne_data <- Rtsne::Rtsne(t(adjusted_counts))
tsne_plot_df <- tsne_data$Y %>%
    data.frame() %>%
    magrittr::set_colnames(c("TSNE1", "TSNE2")) %>%
    mutate("sampleid" = colnames(adjusted_counts)) %>%
    left_join(., mdata)

tsne_plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = tsne_plot_df,
        x_variable = "TSNE1",
        y_variable = "TSNE2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = tsne_plots, ncol = 3)
final_plot
```


### Do ORA to see what PC1 and PC2 correspond to
```{r}
# this is needed for the enrichment analysis that we are doing.
hs_msigdb_df <- msigdbr::msigdbr(species = "Homo sapiens")

gene_sets <- hs_msigdb_df %>%
    filter(gs_subcat == "GO:BP" | gs_cat == "H" | gs_subcat == "CP:KEGG") %>%
    select(gs_name, gene_symbol)


# Extract the genes which most influence PC1 and PC2 and see if they are enriched in any biological domain:
ORA_results <- parallel::mclapply(1:4, function(z) {
    print(paste("Running ORA for PC", z))
    # extract top 300 genes which most influence PC
    top_loadings_gene_names <- pca_data$xvars[order(
        abs(pca_data$loadings[, paste0("PC", z)]),
        decreasing = T
    )[1:300]]

    enrichment_result <- enricher(
        gene = top_loadings_gene_names,
        universe = rownames(dds_filtered),
        minGSSize = 5,
        maxGSSize = 300,
        TERM2GENE = gene_sets
    )

    # for each of the biological domains, grep the top 3 sig. terms
    sig_results <- enrichment_result@result %>%
        mutate(domain = str_extract(Description, "^[^_]*")) %>%
        mutate(ratio_numeric = sapply(.$GeneRatio, function(x) {
            eval(
                str2expression(x)
            )
        })) %>%
        filter(p.adjust < 0.05) %>%
        group_by(domain) %>%
        slice_min(., order_by = p.adjust, n = 3) %>%
        mutate(Description_short = shorten_strings(Description, 30))

    if (nrow(sig_results) > 0) {
        enrichment_plot <- ggplot(sig_results, aes(x = ratio_numeric, y = Description_short, color = p.adjust)) +
            geom_segment(aes(
                x = 0, xend = ratio_numeric, y = Description_short, yend = Description_short
            ), color = "black", linewidth = 1.5) +
            geom_point(size = 5) +
            scale_color_gradient(low = "red", high = "blue") +
            theme_Publication_side() +
            ylab("Description") +
            xlab("Gene Ratio") +
            scale_x_continuous(expand = c(0, 0.01))
    } else {
        enrichment_plot <- ggplot() +
            theme_void()
    }
    return(enrichment_plot)

}, mc.cores=10)
names(ORA_results) <- paste0("PC",1:4)

# maybe not ideal... PC3 and PC4 are decent though (but not a lot of variance)
ORA_results$PC1
ORA_results$PC2
ORA_results$PC3
ORA_results$PC4
```

### Conclusion
Honestly I barely notice any difference between the 2 datasets... I do not think I will remove any samples.

```{r}
sessionInfo()
```