---
title: "RNA seq batch correction"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 150
)
```

```{r, load libraries}
library(DESeq2)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(magrittr)
library(PCAtools)
library(umap)
library(patchwork)
library(sva)
library(httpgd)
library(tidyverse)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


## **Introduction**
For the eQTL mapping, we currently follow the very simple PCA adjustment approach (regress out different number of PCs and ultimately pick the configuration that results in the most eQTLs).
If number is too high, you are also regressing out genotype effects and it will fall. 

Here I want to do 2 things:
1) Check how our data looks like if we also do the PCA approach. 
2) Do a 2 Stage SVA approach like they did in Oli Stegles [peer paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2865505/). 

First regress out known covars, then run SVA on the rest. Results should match the "one step" approach that I am currently using. 



### Use PCA regression

```{r}
dds <- readRDS(here("QC", "results", "dds_final.rds"))
mdata <- colData(dds) %>% data.frame()
norm_data <- assay(vst(dds, blind = TRUE))

# recalculate PCA with adjusted counts
pca_data <- PCAtools::pca(norm_data,
    metadata = colData(dds),
    removeVar = 0.4
)

pca_plot_df <- pca_data$rotated[, 1:50] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:50) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}
```


### Plot PDF
```{r}
# Plot the resulting PCA after regressing PC1 to PC50.
# Each pdf page should be one plot.

# can be done in parallel. Writing to disk/plotting not.
PCA_plots <- pbmcapply::pbmclapply(1:50, function(i) {
    print(i)
    mod <- model.matrix(~group, mdata)
    # Create the formula dynamically
    formula <- as.formula(paste("~", paste0("PC", 1:i, collapse = " + ")))
    mod0 <- model.matrix(formula, pca_plot_df)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

    # adjust the counts using limma
    adjusted_counts <- limma::removeBatchEffect(norm_data,
        covariates = mod0,
        design = mod
    )
    # recalculate PCA with adjusted counts
    pca_data <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds),
        removeVar = 0.4
    )
    pca_plot_df <- pca_data$rotated[, 1:20] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))
    for (z in 1:20) {
        pca_plot_df[, str_interp("PC${z}_explained_var")] <- pca_data$variance[z]
    }

    umap_data <- umap::umap(t(adjusted_counts))
    umap_plot_df <- umap_data$layout %>%
        data.frame() %>%
        magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
        rownames_to_column("sampleid") %>%
        left_join(., mdata)


    columns_to_plot <- c(
        "state", "condition", "group",
        "ethnicity", "age", "batch1id"
    )

    #############
    # PCA PLOTS #
    #############
    plots <- lapply(columns_to_plot, function(x) {
        k = c(1, 2)
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${k[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${k[2]}_explained_var")], 2)
        if (x == "group") {
            tmp_col = c("#fb867d", "#86f88f", "#7a8ff4", "#FA392A", "#0AB517", "#0b218d")
        } else {
            tmp_col = NA
        }
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${k[1]}"),
            y_variable = str_interp("PC${k[2]}"),
            colors = tmp_col,
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${k[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${k[2]}: ${expl_var2}% variance")
        )
        p <- p +
            plot_annotation(
                title = str_interp("PC${i} regression - ${x}"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)

    ##############
    # UMAP PLOTS #
    ##############
    plots <- lapply(columns_to_plot, function(x) {
        if (x == "group") {
            tmp_col = c("#fb867d", "#86f88f", "#7a8ff4", "#FA392A", "#0AB517", "#0b218d")
        } else {
            tmp_col = NA
        }
        p <- density_scatter(
            df = umap_plot_df,
            x_variable = "UMAP1",
            y_variable = "UMAP2",
            colors = tmp_col,
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = "UMAP1",
            ylab = "UMAP2"
        )
        p <- p +
            plot_annotation(
                title = str_interp("PC${i} regression - ${x}"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot_umap <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    return(list(
        "pca_plot" = final_plot,
        "umap_plot" = final_plot_umap
    ))
}, mc.cores = 20)

pdf(here("QC", "results", "imgs", "PC_adjustment", "PCA_adjustment.pdf"), width = 18, height = 11.5)
for (element in PCA_plots){
    print(element[[1]])
    print(element[[2]])
}
dev.off()
```



### Animate the PCA regression
Here, we preserve differences in group, by first regressing it out and then essentially 
running the PC regrssion on the resulting residuals. 
```{r}
library(gganimate)
library(httpgd)
# we have to create one big df that contains information about all PCs
PCA_df_combined <- pbmcapply::pbmclapply(1:50, function(i) {
    print(i)
    mod <- model.matrix(~group, mdata)
    # Create the formula dynamically
    formula <- as.formula(paste("~", paste0("PC", 1:i, collapse = " + ")))
    mod0 <- model.matrix(formula, pca_plot_df)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

    # adjust the counts using limma
    adjusted_counts <- limma::removeBatchEffect(norm_data,
        covariates = mod0,
        design = mod
    )
    # recalculate PCA with adjusted counts
    pca_data_tmp <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds),
        removeVar = 0.4
    )
    pca_plot_df_tmp <- pca_data_tmp$rotated[, 1:2] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid")) %>%
        select(sample_id, PC1, PC2, group)
    for (z in 1:2) {
        pca_plot_df_tmp[, str_interp("PC${z}_explained_var")] <- pca_data_tmp$variance[z]
    }
    pca_plot_df_tmp <- pca_plot_df_tmp %>% mutate(regressionPC = paste0(i))

    # for the animation, it is easier, if the same clusters stay in the same corner.
    # for example M1_ctrl should be top right and M2_stim bottom left.
    # to ensure this, we can multiply PC1 or PC2 *-1 if necessary

    # calculate centriod per group
    centriods <- pca_plot_df_tmp %>%
        group_by(group) %>%
        summarise(
            PC1_mean = mean(PC1),
            PC2_mean = mean(PC2)
        ) %>%
        filter(group %in% c("M1_ctrl", "M2_stim")) %>% 
        column_to_rownames("group")
    
    # if both centriods are positive for M1_ctrl and neg for M2_stim dont do anything
    PC1_multiplier = 1
    PC2_multiplier = 1
    if (centriods["M1_ctrl", 1] < 0) {
        PC1_multiplier = -1
    }
    if (centriods["M1_ctrl", 2] < 0) {
        PC2_multiplier = -1
    }
    pca_plot_df_tmp$PC1 = pca_plot_df_tmp$PC1 * PC1_multiplier
    pca_plot_df_tmp$PC2 = pca_plot_df_tmp$PC2 * PC2_multiplier

    # recalculate centriods to see if it worked
    centriods <- pca_plot_df_tmp %>%
        group_by(group) %>%
        summarise(
            PC1_mean = mean(PC1),
            PC2_mean = mean(PC2)
        ) %>%
        filter(group %in% c("M1_ctrl", "M2_stim")) %>%
        column_to_rownames("group")
    
    if(sum(sign(centriods)[1,]) != 2){
        stop("centriods are wrong")
    }
    return(pca_plot_df_tmp)
}, mc.cores = 25) %>%
    bind_rows()

PCA_df_combined$regressionPC <- factor(PCA_df_combined$regressionPC, levels=1:50)


# the animations look nicer if we do not change axis constantly.
# this means we make sure that all datapoints fall within -30 and 30
dim(PCA_df_combined)
PCA_df_combined <- PCA_df_combined %>%
    filter(PC1 < 40) %>% 
    filter(PC1 > -40) %>% 
    filter(PC2 < 40) %>% 
    filter(PC2 > -40) 
dim(PCA_df_combined)

p <- ggplot(PCA_df_combined, aes(x = PC1, y = PC2, color = group)) +
    geom_point() +
    theme_Publication() +
    guides(color=guide_legend(nrow=2,byrow=TRUE))+
    scale_color_manual(values = c("#fb867d", "#86f88f", "#7a8ff4", "#FA392A", "#0AB517", "#0b218d"))



anim <- p +
    transition_states(regressionPC,
        transition_length = 13,
        state_length = 20,
        wrap = FALSE
    ) +
    labs(title = "Number of regressed PCs: {closest_state}")

anim_save(here("QC", "results", "imgs", "PC_adjustment", "PC_adjusting_preserving_group.gif"),
    animation = anim,
    nframes = 1350,
    fps = 25,
    start_pause = 10,
    end_pause = 30,
    rewind = FALSE,
)
```


Here, we do NOT preserve the differences
```{r}
# we have to create one big df that contains information about all PCs
PCA_df_combined_no_preservation <- pbmcapply::pbmclapply(1:50, function(i) {
    print(i)
    # Create the formula dynamically
    formula <- as.formula(paste("~", paste0("PC", 1:i, collapse = " + ")))
    mod0 <- model.matrix(formula, pca_plot_df)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

    # adjust the counts using limma
    # we do not specify a design which means that biological differences wont be preserved.
    adjusted_counts <- limma::removeBatchEffect(norm_data,
        covariates = mod0,
    )
    # recalculate PCA with adjusted counts
    pca_data_tmp <- PCAtools::pca(adjusted_counts,
        metadata = colData(dds),
        removeVar = 0.4
    )
    pca_plot_df_tmp <- pca_data_tmp$rotated[, 1:2] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid")) %>%
        select(sample_id, PC1, PC2, group)
    for (z in 1:2) {
        pca_plot_df_tmp[, str_interp("PC${z}_explained_var")] <- pca_data_tmp$variance[z]
    }
    pca_plot_df_tmp <- pca_plot_df_tmp %>% mutate(regressionPC = paste0(i))

    # for the animation, it is easier, if the same clusters stay in the same corner.
    # for example M1_ctrl should be top right and M2_stim bottom left.
    # to ensure this, we can multiply PC1 or PC2 *-1 if necessary

    # calculate centriod per group
    centriods <- pca_plot_df_tmp %>%
        group_by(group) %>%
        summarise(
            PC1_mean = mean(PC1),
            PC2_mean = mean(PC2)
        ) %>%
        filter(group %in% c("M1_ctrl", "M2_stim")) %>% 
        column_to_rownames("group")
    
    # if both centriods are positive for M1_ctrl and neg for M2_stim dont do anything
    PC1_multiplier = 1
    PC2_multiplier = 1
    if (centriods["M1_ctrl", 1] < 0) {
        PC1_multiplier = -1
    }
    if (centriods["M1_ctrl", 2] < 0) {
        PC2_multiplier = -1
    }
    pca_plot_df_tmp$PC1 = pca_plot_df_tmp$PC1 * PC1_multiplier
    pca_plot_df_tmp$PC2 = pca_plot_df_tmp$PC2 * PC2_multiplier

    # recalculate centriods to see if it worked
    centriods <- pca_plot_df_tmp %>%
        group_by(group) %>%
        summarise(
            PC1_mean = mean(PC1),
            PC2_mean = mean(PC2)
        ) %>%
        filter(group %in% c("M1_ctrl", "M2_stim")) %>%
        column_to_rownames("group")
    
    if(sum(sign(centriods)[1,]) != 2){
        stop("centriods are wrong")
    }
    return(pca_plot_df_tmp)
}, mc.cores = 25) %>%
    bind_rows()
PCA_df_combined_no_preservation$regressionPC <- factor(PCA_df_combined_no_preservation$regressionPC, levels=1:50)


# the animations look nicer if we do not change axis constantly.
# this means we make sure that all datapoints fall within -30 and 30
dim(PCA_df_combined_no_preservation)
PCA_df_combined_no_preservation <- PCA_df_combined_no_preservation %>%
    filter(PC1 < 40) %>% 
    filter(PC1 > -40) %>% 
    filter(PC2 < 40) %>% 
    filter(PC2 > -40) 
dim(PCA_df_combined_no_preservation)

p <- ggplot(PCA_df_combined_no_preservation, aes(x = PC1, y = PC2, color = group)) +
    geom_point() +
    theme_Publication() +
    guides(color=guide_legend(nrow=2,byrow=TRUE))+
    scale_color_manual(values = c("#fb867d", "#86f88f", "#7a8ff4", "#FA392A", "#0AB517", "#0b218d"))


PC1_expl_var <- unique(PCA_df_combined_no_preservation$PC1_explained_var)
PC2_expl_var <- unique(PCA_df_combined_no_preservation$PC2_explained_var)
anim <- p +
    transition_states(regressionPC,
        transition_length = 13,
        state_length = 20,
        wrap = FALSE
    ) +
    labs(
        title = "Number of regressed PCs: {closest_state}",
        xlab = paste0("PC1:",PC1_expl_var['{closest_state}']  ,"% explained variance"),
        ylab = paste0("PC2:",PC2_expl_var['{closest_state}']  ,"% explained variance"),
        )
        
anim_save(here("QC", "results", "imgs", "PC_adjustment", "PC_adjusting_no_preserving_group.gif"),
    animation = anim,
    nframes = 1350,
    fps = 25,
    start_pause = 10,
    end_pause = 30,
    rewind = FALSE,
)



# anim <- p +
#     transition_states(regressionPC,
#         transition_length = 300,
#         state_length = 400,
#         wrap = FALSE
#     ) +
#     labs(title = "Number of regressed PCs: {closest_state}")



```


### Do the 2 step SVA regression
```{r}
# step one is to regress out known covars
norm_data_adj <- limma::removeBatchEffect(norm_data,
    batch = dds$line,
    group = dds$group
)


# just 9 values, shouldnt be an issue.
sum(norm_data_adj<0)
norm_data_adj[norm_data_adj < 0] = 0

# Now run the "default" sva on the residuals/adjusted expression
mod <- model.matrix(~ group, colData(dds))
mod0 <- model.matrix(~1, colData(dds))
svseq_two_step <- svaseq(norm_data_adj, mod, mod0, n.sv = 5)

# load our approach 
svseq_one_step <- readRDS(here("QC", "results", "sva_5_final.rds"))

# completely different. 
cor(svseq_two_step$sv[, 1], svseq_one_step$sv[, 1])













###########
# HEATMAP #
###########
SVs <- svseq_two_step$sv %>% magrittr::set_colnames(paste0("SV", 1:5))
mdata_two_step <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)


covar_list = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize", "line")
correlation_SVs <- covar_correlation(
    df = mdata_two_step,
    covars = covar_list,
    components = paste0("SV", 1:5)
)

correlation_SVs$heatmap

adjusted_two_step <- limma::removeBatchEffect(
    norm_data,
    batch = mdata_two_step$line,
    covariates = model.matrix(~SV1+SV2+SV3+SV4+SV5, mdata_two_step)[,-1],
    group = mdata_two_step$group
)

# recalculate PCA with adjusted counts
pca_data <- PCAtools::pca(adjusted_two_step,
    metadata = mdata_two_step,
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (k in 1:20) {
    pca_plot_df[, str_interp("PC${k}_explained_var")] <- pca_data$variance[k]
}
k = c(1, 2)
columns_to_plot <- c("state", "condition", "group", "ethnicity", "age", "batch1id")
plots <- lapply(columns_to_plot, function(x) {
    expl_var1 <- round(pca_plot_df[1, str_interp("PC${k[1]}_explained_var")], 2)
    expl_var2 <- round(pca_plot_df[1, str_interp("PC${k[2]}_explained_var")], 2)
    if (x == "group") {
        tmp_col = c("#fb867d", "#86f88f", "#7a8ff4", "#FA392A", "#0AB517", "#0b218d")
    } else {
        tmp_col = NA
    }   
    p <- density_scatter(
        df = pca_plot_df,
        x_variable = str_interp("PC${k[1]}"),
        y_variable = str_interp("PC${k[2]}"),
        colors = tmp_col,
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
        xlab = str_interp("PC${k[1]}: ${expl_var1}% variance"),
        ylab = str_interp("PC${k[2]}: ${expl_var2}% variance")
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)

# this looks actually very similar to the plot using our approach
final_plot


# I think that our SV association plot also makes sense actually. 
# We have regressed out effects of sex, ethnicity, age, batch and so on by regressing line
# consequently the SVs CANT pick this up. Instead they are picking stuff up which is not in our meta data.
# in that way, it is actually a positive control that we dont observe correlations. 
SVs <- svseq_one_step$sv %>% magrittr::set_colnames(paste0("SV", 1:5))
mdata_one_step <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)


covar_list = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize", "line")
correlation_SVs <- covar_correlation(
    df = mdata_one_step,
    covars = covar_list,
    components = paste0("SV", 1:5)
)
correlation_SVs$heatmap



```



### Quick check of the used genes

```{r}
# I noticed a lot of pseudo genes, see if this is actually the case
library(biomaRt)
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
attributes <- c("external_gene_name", "gene_biotype")
biomart_results <- getBM(attributes = attributes,
                 filters = "external_gene_name",
                 values = rownames(norm_data),
                 mart = ensembl)

dim(norm_data)
dim(biomart_results)
head(biomart_results)

biomart_results$gene_biotype %>% table() %>% sort()

# genes which are protein coding or annotated lncRNA:
13908+1043

# 3k genes are pseudogenes and the like... check the actual input data of the pipeline why there are so many
nrow(norm_data)

rownames(norm_data)[!rownames(norm_data) %in% biomart_results$external_gene_name]



```

```{r}
sessionInfo()
```





