---
title: "HipSci cell lines name check"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 350
)
```

## **Introduction**
The project uses HipSci cell lines. A lot of information about them is publically available. 
Here we see what kind of information we can find about the HipSci lines that we are using in 
the oxLDL project.


```{r, load libraries}
library(tidyverse)
library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```


## **Load oxLDL mdata and HipSci information**
```{r, load in data}
# important: NAs are encoded as strings ("NA")
mdata_oxLDL <- readxl::read_excel(here("data", "mdata_google_spreadsheet.xlsx"),
    sheet = "RNA Sequencing samples"
) %>% data.frame()
mdata_hipsci <- data.table::fread(
    here("data", "oxLDL_data", "HipSci", "HipSci_cell_lines.tsv")
) %>% data.frame()

mdata_hipsci_donor <- data.table::fread(
    here("data", "oxLDL_data", "HipSci", "HipSci_donors.tsv")
) %>% data.frame()

# in our data HipSci cell line is encoded in the "line" column.
# I am unsure what the number means, but I do not think that it is really relevant so remove it.
# -> each line is uniquely identifed by 4 letters
oxLDL_line_id <- substr(mdata_oxLDL$line, 1, 4) %>% unique()

# in the hipsci data, the line information is encoded in the name.
# we will again remove the number
hipsci_line_id <- gsub(".*-(\\w+)", "\\1", mdata_hipsci$Name) %>%
    str_remove_all(., "_") %>%
    str_remove_all(., "\\d") %>%
    unique()

mdata_hipsci$id <- gsub(".*-(\\w+)", "\\1", mdata_hipsci$Name) %>%
    str_remove_all(., "_") %>%
    str_remove_all(., "\\d")

hipsci_donor_id <- mdata_hipsci_donor$Name %>% str_remove_all(., "HPSI-") %>% unique()


# go through our cell lines and see if we can identify a match in the hipsci data
overlap_cell_line <- sapply(oxLDL_line_id, function(id) {
    return(any(grepl(id, hipsci_line_id)))
})
# 3 of our samples we do not find in the hipsci data
table(overlap_cell_line)
overlap_cell_line[overlap_cell_line==FALSE]
# google also does not return any results for these 3 cell lines...
# also cant find them here: /rds/project/jmmh2/rds-jmmh2-results/private/oxLDL/HipSci/HipSci_sample_biosamples.json

# no idea what is "wrong" with these samples... 
mdata %>% filter(grepl("jacg", line))
mdata %>% filter(grepl("guif", line))
mdata %>% filter(grepl("guio", line))


mdata_hipsci %>%
    filter(id %in% oxLDL_line_id) %>%
    select(id, Donor.Consent) %>%
    distinct(., .keep_all = TRUE) %>%
    pull(Donor.Consent) %>% 
    table()


# if we check for overlap with donor, there are 3 additional samples for which
# we do not find a match:
# foqj, yuyn and ffdl
# if we google them though, we can see that they show up for hipsci
overlap_donor <- sapply(oxLDL_line_id, function(id) {
    return(any(grepl(id, hipsci_donor_id)))
})
table(overlap_donor)
overlap_donor[overlap_donor==FALSE]
mdata_oxLDL %>% filter(grepl("foqj", line))
```



## **Overlap with downloaded genotype data**
We have also downloaded some genotype information. Check for overlap there.
```{r}
genotype_mdata <- data.table::fread(
    here("data", "oxLDL_data", "HipSci", "genotype_array_data", "filereport_analysis_PRJEB11750_tsv.txt")
) %>% data.frame()

ftp_files <- genotype_mdata$submitted_ftp %>% str_split(., ";") %>% unlist()
# remove all strings that contain ped
ftp_files <- ftp_files[!grepl("\\.ped", ftp_files)]


# in total we have 695 unique ids. if we remove the number this reduces to 242
genotype_id <- gsub('.*-(\\w+)\\.we.*', "\\1", ftp_files) %>% str_remove_all(., "_\\d") %>% unique()

overlap_genotype <- sapply(oxLDL_line_id, function(id) {
    return(any(grepl(id, genotype_id)))
})
table(overlap_genotype)
overlap_genotype[overlap_genotype==FALSE]
```
