---
title: "RNA seq qc"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 350
)
```

## **Introduction**
The [nf-core RNAseq pipeline](https://github.com/nf-core/rnaseq) already does a very extensive set of quality control steps 
based on the read level (for example # Duplicated reads, GC content, 
mapping rate, read distribution, …). 
In this script, we will do some qulaity control steps based on the 
quantified counts (using [STAR](https://github.com/alexdobin/STAR) followed by [salmon](https://combine-lab.github.io/salmon/)).
 We will also make a few plots to explore the metadata


```{r, load libraries}
library(DESeq2)
library(magrittr)
library(PCAtools)
library(tidyverse)
library(umap)
library(patchwork)
library(httpgd)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```



```{r}
plot_column_table <- function(input_vector, include_n=FALSE) {
  # Convert input vector to data frame
  data_df <- data.frame(values = input_vector)
  
  # Create the ggplot barplot
  plot <- ggplot(data_df, aes(x = values)) +
      geom_bar(fill = "#0f8fc2", color = "black") +
      labs(title = "Frequency of Elements", x = "Elements", y = "Frequency") +
      scale_y_continuous(expand = c(0.01, 0)) +
      theme_Publication()
  
  if(include_n){
    data_summary <- data_df %>%
        group_by(values) %>%
        summarise(count = n())
    
    max_count <- max(data_summary$count)
    y_max <- max_count*1.25   # Increase by 20% to add a buffer

    plot <- plot +
        geom_text(
            data = data_summary, aes(label = count, y = count),
            vjust = -0.5, color = "black", size = 4
        )+
        scale_y_continuous(expand = c(0.01, 0), limits=c(0,y_max))
  }
  return(plot)
}
```
## **Description of metadata**
Few observations: 
(a) A few important confounders (e.g. `age`, `ethnicity`) are missing for some samples 
(b) There seem to be a few technical replicates (cell lines that were used more than 6 times) 
(c) MM_oxLDL7159504 seems to contain conflicting information. 
oxLDL treated according to column `oxLDL` but control according to column `sampeltype`.

```{r}
mdata <- readxl::read_excel(here("data", "mdata_google_spreadsheet.xlsx"),
    sheet = "RNA Sequencing samples"
) %>% data.frame()

submission_mdata <- data.table::fread(here("data", "oxLDL_data", "RNA_seq", "macromap_samplesheet.csv"),
    data.table = FALSE
)
mdata$ethnicity <- ifelse(mdata$ethnicity == "White - White other", "White - Other", mdata$ethnicity)
mdata$ethnicity <- ifelse(mdata$ethnicity == "Asian /Asian British- Indian", "Asian British- Indian", mdata$ethnicity)
factor_columns <- c("state", "oxldl", "set", "ethnicity", "pool", "age", "batch1id", "oxldlset")
mdata[factor_columns] <- lapply(mdata[factor_columns], factor)

stopifnot(all(mdata$sampleid %in% submission_mdata$sample))
stopifnot(n_distinct(mdata$sampleid) == nrow(mdata))
mdata %>%
    filter(grepl("_0", sampletype)) %>%
    filter(oxldl != 0)
mdata %>%
    filter(grepl("_oxLDL", sampletype)) %>%
    filter(oxldl != 100)

age <- plot_column_table(mdata$age, include_n = T) +
    ggtitle("Age distribution") +
    xlab("Age") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

rna_pool <- plot_column_table(mdata$pool %>%
    str_remove_all("RNA ") %>%
    factor(., levels = c("pilot", as.character(1:17))), include_n=TRUE) +
    ggtitle("RNA pool distribution") +
    xlab("RNA pool")


type <- plot_column_table(mdata$state, include_n = T) +
    ggtitle("Macrophage type distribution") +
    xlab("Macrophage type")

# some lines have more than the expected number (6) of entries. Take a look at them
# we might have to remove one of them because they are technical replicates and not biological ones.
lines <- plot_column_table(mdata$line) +
    ggtitle("Cell line distribution") +
    xlab("Cell lines") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

# simple technical replicates sometimes, even in the same RNA pool and batch
# for example MM_oxLDL9752052 and MM_oxLDL9752058
mdata %>%
    group_by(line) %>%
    mutate(n_lines = n()) %>%
    filter(n_lines > 6) %>%
    arrange(line) 

condition <- plot_column_table(ifelse(mdata$oxldl == "100", "stimulated", "unstimulated"),
    include_n = T
)+
    ggtitle("Condition distribution") +
    xlab("Condition")


oxLDL_set <- plot_column_table(factor(mdata$oxldlset))+
    ggtitle("oxLDL set distribution") +
    xlab("oxLDL set") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

sex <- plot_column_table(mdata$set, include_n = TRUE) +
    ggtitle("Sex distribution") +
    xlab("Sex")

ethnicity_vector <- mdata %>%
    distinct(line, .keep_all = TRUE) %>%
    pull(ethnicity)

ethnicity <- plot_column_table(
    factor(ethnicity_vector,
        levels = c("White - White British", "White - Other", "Asian British- Indian", "NA")
    ),
    include_n = TRUE
) +
    ggtitle("Ethnicity") +
    xlab("Ethnicity") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

batches <- plot_column_table(as.character(mdata$batch1id), include_n = TRUE)+
    ggtitle("Batches distribution") +
    xlab("Batch")+
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

    # combine the plots
combined_plot <- wrap_plots(
    age, sex, ethnicity,
    lines, oxLDL_set, batches,
    type, condition,
    ncol = 3,
    byrow = TRUE
)
ggsave(combined_plot,
    filename = here("QC", "results", "imgs", "mdata_overview.png"), 
    dpi = 350, width = 18, height = 8
)
```

## **Count data quality control**
Read in the count data generated by the pipeline. They were slightly modified using the 
script preprocessing/scripts/unify_geneID_to_geneName.rmd


Clearly strong batch effects. Main culprits are:
(a) batch1id
(b) pool (overlaps with batch1id)
(c) age

Our variables of interest seem to not associate with any PCs.
 
```{r}
raw_counts <- data.table::fread(
    here("preprocessing", "results", "gene_counts_length_scaled_filtered.tsv")
) %>% data.frame()

raw_counts_matrix <- as.matrix(raw_counts[,2:ncol(raw_counts)])
rownames(raw_counts_matrix) <- raw_counts$gene_name

# make sure mdata order matches the column names of raw_counts
mdata <- mdata[match(colnames(raw_counts_matrix), mdata$sampleid), ]
rownames(mdata) <- mdata$sampleid

dds0 <- DESeqDataSetFromMatrix(round(raw_counts_matrix),
    colData = mdata,
    design = ~1
)
# remove some lowly expressed genes
keep <- edgeR::filterByExpr(counts(dds0), group = mdata$sampletype)
dds0 <- dds0[keep,]
vsd_data <- vst(dds0)
norm_data <- assay(vsd_data)
```



```{r}
#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(norm_data,
    metadata = colData(dds0),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

##################
# CALCULATE UMAP #
##################
umap_data <- umap(t(norm_data))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>% 
    left_join(., mdata)

###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "oxldl", "oxldlset", "set", "ethnicity", "age", "batch1id"),
    components = paste0("PC", 1:15)
)

png(here("QC", "results", "imgs", "association_heatmap.png"), res=350, width=9, height=5.5, units="in")
association_output$heatmap
dev.off()

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "oxldl", "set", "ethnicity", "pool", "age", "batch1id")

PC_pairings <- lapply(1:7, function(x){
    c(2*x-1,2*x)
})

for (i in PC_pairings) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_pca(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )


        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "QC", "results", "imgs", "SVA_10",
            str_interp("PCA_all_covariates_PC${i[1]}_PC${i[2]}_corrected.png")
        ),
        dpi = 350,
        width = 20,
        height = 15,
    )
}

################
# BOXPLOTS PCA #
################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = pca_plot_df,
        covar = x,
        PC = paste0("PC", 1:4)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "PCA_all_covariates_boxplots.png"),
    dpi = 350,
    width = 20,
    height = 15,
)

##############
# UMAP PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_pca(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "UMAP_all_covariates.png"),
    dpi = 350,
    width = 20,
    height = 15,
)

################
# BOXPLOTS UMAP #
################
plots <- lapply(columns_to_plot, function(x) {
    p <- plot_categorical_covar(
        df = umap_plot_df,
        covar = x,
        PC = paste0("UMAP", 1:2)
    ) %>% wrap_plots(., ncol = 2) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size=5, face="bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "UMAP_all_covariates_boxplots.png"),
    dpi = 350,
    width = 20,
    height = 15,
)
```


## **Batch correction**

### SVA
Try out [sva](). 
`svaseq` expects normalized input data. Otherwise the first surrogate variable is often the lirbary size.
See [Jeff Leeks reply](https://support.bioconductor.org/p/76381/). We take normalized DESeq2 counts.
Internally first step of `svaseq` is log transformation so important that our normalized counts are not log transformed.

```{r}
library(sva)
# sva package also provides a function to estimate the number of SVs we should use.
# lets try the function:

# the documentation for this function is sparse, but I think it expects log transformed counts
# that were normalized with respect to library size
# at least that is the data that is used in the example of the function so we will do the same. 
# this runs very long. 
mod  <- model.matrix(~ sampletype, colData(dds0))
mod0 <- model.matrix(~1, colData(dds0))


# this returns 54 which is kind of is definitely way too many...
#estimated_n_SVs = num.sv(norm_data, mod = mod, method = "be", B = 30)
# this return 3 which seems too few...
#estimated_n_SVs_leek = num.sv(norm_data, mod=mod, method= "leek")


dds0 <- estimateSizeFactors(dds0)
norm_counts  <- counts(dds0, normalized = TRUE)

mod  <- model.matrix(~ sampletype, colData(dds0))
mod0 <- model.matrix(~1, colData(dds0))

# important: The number of SVs matters.
# I.e. if you try 2 runs with 10 or 15, the first 10 of the run with 15 wont be the same as the other run.
#svseq <- svaseq(norm_counts, mod, mod0, n.sv = 10)
#saveRDS(svseq, here("QC", "results", "sva_10SV.rds"))
svseq <- readRDS(here("QC", "results", "sva_15SV.rds"))
```


## **Try and find out what the SVs correspond to**

```{r}
# try and find out what the SVs correspond to. For this we can use the same functions as above
# I assume that the order in svseq does not change, because that would just be random
SVs <- svseq$sv
colnames(SVs) <- paste0("SV",1:15)
mdata <- cbind(mdata %>% select(-contains("SV")), data.frame(SVs))
covar_list = c("pool", "state", "oxldl", "oxldlset", "set", "ethnicity", "age", "batch1id")
correlation_SVs <- covar_correlation(
    df = mdata,
    covars = covar_list,
    components =paste0("SV",1:15)
)
png(here("QC", "results", "imgs", "SVA_15", "association_heatmap.png"), res=350, width=9, height=5.5, units="in")
correlation_SVs$heatmap
dev.off()

# check for association
for (x in covar_list) {
    print(x)
    p <- plot_categorical_covar(
        df = mdata,
        covar = x,
        PC = paste0("SV", 1:15),
        bonferroni_correction = TRUE
    )
    p <- lapply(p, function(k) {
        k <- k + geom_hline(yintercept = 0, linetype = "dashed")
    })

    p <- p %>% wrap_plots(., ncol = 5) &
        theme(
            axis.text.x = element_blank(),
            axis.text.y = element_text(size = 5, face = "bold")
        )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    ggsave(
        plot = p,
        filename = here("QC", "results", "imgs", "SVA_15", str_interp("SV_associations_${x}.png")),
        dpi = 350,
        width = 15,
        height = 10,
    )
}


## ALSO CREATE A PC PLOT WHERE INSTEAD OF PLOTTING PCS WE PLOT SV1 and 2 AS THE AXIS -> SEE IF THIS BETTER
## SEPARATES SHIT






```

## Batch corrected PCA 
Seems to work very well. We do see separation of Macrophage subtypes now and oxLDL status, but this is also what we would expect 
given that we specified that this is the comparison of interest. If we just use 10 SVs, we can still observe a clear clustering of variables according to batch. 
This is drastically reduced when we increase the number of SVs. The exampled variance also goes down if we increase number of SVs.


```{r}
################
# USING 10 SVs #
################
svseq <- readRDS(here("QC", "results", "sva_10SV.rds"))
SVs <- svseq$sv
colnames(SVs) <- paste0("SV",1:10)
colData(dds0) <- cbind(colData(dds0), SVs)

# use limmas removeBatchEffects to emulate what the linear model will eventually do.
norm_data <- assay(vsd_data)
mod  <- model.matrix(~ sampletype, colData(dds0))
mod0 <- model.matrix(
    ~ SV1 + SV2 + SV3 + SV4 + SV5 + SV6 + SV7 + SV8 + SV9 + SV10,
    colData(dds0)
)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/


adjusted_counts <- limma::removeBatchEffect(norm_data,
    covariates = mod0, 
    design = mod
)

# recalculate PCA with adjusted counts
pca_data <- PCAtools::pca(adjusted_counts,
    metadata = colData(dds0),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

##################
# CALCULATE UMAP #
##################
umap_data <- umap(t(adjusted_counts))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>% 
    left_join(., mdata)

###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "oxldl", "oxldlset", "set", "ethnicity", "age", "batch1id"),
    components = paste0("PC", 1:10)
)

png(here("QC", "results", "imgs", "SVA_10", "association_heatmap_corrected.png"),
    res = 350, width = 9, height = 5.5, units = "in"
)
association_output$heatmap
dev.off()

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "oxldl", "set", "ethnicity", "pool", "age", "batch1id")

#create PC intervals 
PC_pairings <- lapply(1:7, function(x){
    c(2*x-1,2*x)
})

for (i in PC_pairings) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_pca(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )


        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "QC", "results", "imgs", "SVA_10",
            str_interp("PCA_all_covariates_PC${i[1]}_PC${i[2]}_corrected.png")
        ),
        dpi = 350,
        width = 20,
        height = 15,
    )
}

##############
# UMAP PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_pca(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "SVA_10","UMAP_all_covariates_corrected.png"),
    dpi = 350,
    width = 20,
    height = 15,
)

# DO CO OCCURANCE PLOTS 
# DO ALSO PCA WHEN I REMOVE THE ACTUAL CONFOUNDERS WITHOUT ANY SURROGATES
# maybe write a function which does all these plots? Will probably increase reability? 
```


## **Batch correction using mdata**
One very big downside of this method is, that all samples that have NA in any of the variables that are important
for batch correction have to be dropped, because we simply do not to which group they belong.

Nonetheless, try out how it looks like.

```{r}
# we will adjust for the following variables:
covars_to_adjust <- c("age", "ethnicity", "batch1id", "pool", "oxldlset")

# first we have to remove all samples from the dataset that have NA in any of the rows
keep_sampleids <- colData(dds0) %>%
    data.frame() %>%
    filter(!if_any(c(age, ethnicity, batch1id, pool, oxldlset), ~ . == "NA")) %>% pull(sampleid)

dds_filtered <- dds0[, dds0$sampleid %in% keep_sampleids]

# refactor the columns so NAs get removed
factor_columns <- covars_to_adjust
colData(dds_filtered)[factor_columns] <- lapply(colData(dds_filtered)[covars_to_adjust], factor)
# use limmas removeBatchEffects to emulate what the linear model will eventually do.
norm_data <- assay(vst(dds_filtered, blind = TRUE))
mod <- model.matrix(~sampletype, colData(dds_filtered))


# if we include pool, or oxldl in the model matrix, we introduce linear dependency
mod0 <- model.matrix(~ batch1id + ethnicity + age, colData(dds_filtered))[, -1]
dim(mod0)
qr(mod0)$rank
adjusted_counts <- limma::removeBatchEffect(norm_data,
    covariates = mod0, 
    design = mod
)
 
# recalculate PCA with adjusted counts
pca_data <- PCAtools::pca(adjusted_counts,
    metadata = colData(dds_filtered),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

##################
# CALCULATE UMAP #
##################
umap_data <- umap(t(adjusted_counts))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>% 
    left_join(., mdata)

###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "oxldl", "oxldlset", "set", "ethnicity", "age", "batch1id"),
    components = paste0("PC", 1:15)
)

png(here("QC", "results", "imgs", "mdata_adjustment", "association_heatmap_corrected.png"),
    res = 350, width = 9, height = 5.5, units = "in"
)
association_output$heatmap
dev.off()

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "oxldl", "set", "ethnicity", "pool", "age", "batch1id")

plots <- lapply(columns_to_plot, function(x) {
    p <- density_pca(
        df = pca_plot_df,
        x_variable = "PC1",
        y_variable = "PC2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
        xlab = paste0("PC1: ", round(pca_plot_df$PC1_explained_var[1], 2), "% variance"),
        ylab = paste0("PC2: ", round(pca_plot_df$PC2_explained_var[1], 2), "% variance")
    )

    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})
final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "mdata_adjustment","PCA_all_covariates_PC1_PC2_corrected.png"),
    dpi = 350,
    width = 20,
    height = 15,
)

##############
# UMAP PLOTS #
##############
plots <- lapply(columns_to_plot, function(x) {
    p <- density_pca(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = plots, ncol= 3)
ggsave(
    plot = final_plot,
    filename = here("QC", "results", "imgs", "mdata_adjustment","UMAP_all_covariates_corrected.png"),
    dpi = 350,
    width = 20,
    height = 15,
)
```









However, RUVSeq does have steps to deal with sequencing depth, see vignette section 2.1, the fourth code chunk. So if you are following the RUVSeq workflow including that step you can provide original counts, not scaled counts.
https://support.bioconductor.org/p/9137741/



# This is the vignette for sva essentially

```{r, eval=FALSE}
library(zebrafishRNASeq)
data(zfGenes)
filter = apply(zfGenes, 1, function(x) length(x[x>5])>=2)
filtered = zfGenes[filter,]
genes = rownames(filtered)[grep("^ENS", rownames(filtered))]
controls = grepl("^ERCC", rownames(filtered))
spikes =  rownames(filtered)[grep("^ERCC", rownames(filtered))]
group = as.factor(rep(c("Ctl", "Trt"), each=3))
set = EDASeq::newSeqExpressionSet(as.matrix(filtered),
    phenoData = data.frame(group, row.names = colnames(filtered))
)
dat0 = counts(set)

## Set null and alternative models (ignore batch)
mod1 = model.matrix(~group)
mod0 = cbind(mod1[,1])

## Estimate batch with svaseq (unsupervised)
batch_unsup_sva = svaseq(dat0,mod1,mod0,n.sv=1)$sv
batch_unsup_sva_3 = svaseq(dat0,mod1,mod0,n.sv=3)$sv

batch_unsup_sva[, 1] == batch_unsup_sva_3[, 1]


```

