---
title: "Generate slides for powerpoint recapping QC"
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 300
)
```

## **Introduction**
This is just a concise scripts to generate the plots for the QC ppt.
For more extensive explanations, see the respective scripts

```{r, load libraries}
library(DESeq2)
library(magrittr)
library(PCAtools)
library(tidyverse)
library(umap)
library(patchwork)
library(httpgd)

library(here)
source(here("plot_theme.R"))
source(here("helper_functions.R"))
```

### Metadata before QC
Original script: QC/scripts/01_initial_QC.rmd

```{r, res = 150, width = 14, height = 8}
mdata <- readRDS(here("QC", "results", "mdata_clean.rds"))

age <- plot_column_table(
    mdata %>%
        distinct(line, .keep_all = T) %>%
        pull(age),
    include_n = T
) +
    ggtitle("Age distribution") +
    xlab("Age") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))


rna_pool <- plot_column_table(mdata$pool %>%
    str_remove_all("RNA ") %>%
    factor(., levels = c("pilot", as.character(1:17))), include_n=TRUE) +
    ggtitle("RNA pool distribution") +
    xlab("RNA pool")


state <- plot_column_table(mdata$state, include_n = T) +
    ggtitle("Macrophage type distribution") +
    xlab("Macrophage type")

# some lines have more than the expected number (6) of entries. Take a look at them
# we might have to remove one of them because they are technical replicates and not biological ones.
lines <- plot_column_table(mdata$line) +
    ggtitle("Cell line distribution") +
    xlab("Cell lines") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

condition <- plot_column_table(mdata$condition,
    include_n = T
) +
    ggtitle("Condition distribution") +
    xlab("Condition")


oxLDL_set <- plot_column_table(factor(mdata$oxldlset))+
    ggtitle("oxLDL set distribution") +
    xlab("oxLDL set") +
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

sex <- plot_column_table(
    mdata %>%
        distinct(line, .keep_all = T) %>%
        pull(sex),
    include_n = T
) +
    ggtitle("Sex distribution") +
    xlab("Sex")

ethnicity <- plot_column_table(mdata$ethnicity,
    include_n = TRUE
) +
    ggtitle("Ethnicity") +
    xlab("Ethnicity") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))

batches <- plot_column_table(as.character(mdata$batch1id), include_n = TRUE)+
    ggtitle("Batches distribution") +
    xlab("Batch")+
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )

# combine the plots
combined_plot <- wrap_plots(
    age, sex, ethnicity,
    lines, oxLDL_set, batches,
    state, condition,
    ncol = 3,
    byrow = TRUE
)
ggsave(
    plot = combined_plot,
    filename = here("ppt_scripts", "imgs","mdata_overview.png"),
    dpi = 350, width = 13, height = 8
)
combined_plot
```


## **PCA to remove the libsize outliers**

```{r}
# load the DESeq2 object
dds <- readRDS(here("QC", "results", "dds_original.rds"))
norm_data <- assay(vst(dds, blind = TRUE))

#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(norm_data,
    metadata = colData(dds),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

# Standard PCA plot to highlight the libsize outliers
p <- density_scatter(
    df = pca_plot_df,
    x_variable = str_interp("PC1"),
    y_variable = str_interp("PC2"),
    color_by = "condition",
    density_plot_ratio = 0.15,
    pt_size = 1.2,
    xlab = str_interp("PC1: ${signif(unique(pca_plot_df$PC1_explained_var), digits = 3)}% variance"),
    ylab = str_interp("PC2: ${signif(unique(pca_plot_df$PC2_explained_var), digits = 3)}% variance")
)
p <- p +
    plot_annotation(
        title = "condition",
        theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
    )

ggsave(
    plot = p,
    filename = here("ppt_scripts","imgs", "inital_PCA.png"),
    dpi = 350, width = 4, height = 3,
    scale=2
)

# highlight outliers.
# we cant rescue them, so remove them.
p <- density_scatter(
    df = pca_plot_df %>% mutate("Library size" = ifelse(libsize>15000, "> 15k", "< 15k")),
    x_variable = str_interp("PC1"),
    y_variable = str_interp("PC2"),
    color_by = "Library size",
    density_plot_ratio = 0.15,
    pt_size = 1.2,
    xlab = str_interp("PC1: ${signif(unique(pca_plot_df$PC1_explained_var), digits = 3)}% variance"),
    ylab = str_interp("PC2: ${signif(unique(pca_plot_df$PC2_explained_var), digits = 3)}% variance")
)
p <- p +
    plot_annotation(
        title = "Library size",
        theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
    )

ggsave(
    plot = p,
    filename = here("ppt_scripts", "imgs","libsize_outlier_PCA.png"),
    dpi = 350, width = 4, height = 3,
    scale=2
)
```

## **PCA to highlight batch effects**

```{r}
# this object has the tiny libsize outliers removed
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))

#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(assay(vst(dds, blind=TRUE)),
    metadata = colData(dds),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")

PC_pairings <- lapply(1:3, function(x){
    c(2*x-1,2*x)
})
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
for (i in PC_pairings) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )


        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "ppt_scripts", "imgs",str_interp("PCA_covariates_PC${i[1]}_PC${i[2]}_uncorrected.png")
        ),
        dpi = 350,
        width = 18,
        height = 11.5,
    )
}

################
# BOXPLOTS PCA #
################
p <- plot_categorical_covar(
    df = pca_plot_df,
    covar = "batch1id",
    PC = paste0("PC", 1:4)
) %>% wrap_plots(., ncol = 2) &
    theme(
        axis.text.x = element_blank(),
        axis.text.y = element_text(size=5, face="bold")
    )
p <- p +
    plot_annotation(
        title = "Batch",
        theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
    )

ggsave(
    plot = p,
    filename = here("ppt_scripts", "imgs", "PCA_batch_boxplots.png"),
    dpi = 350,
    width = 10,
    height = 9,
)


#########
#  UMAP #
#########
norm_data <- assay(vst(dds, blind = TRUE))
umap_data <- umap(t(norm_data))

umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>%
    left_join(., colData(dds) %>% data.frame())

columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
ggsave(
    plot = final_plot,
    filename = here(
        "ppt_scripts", "imgs", str_interp("UMAP_covariates_uncorrected.png")
    ),
    dpi = 350,
    width = 18,
    height = 11.5,
)


###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "line"),
    components = paste0("PC", 1:15)
)

png(here("ppt_scripts","imgs", "association_heatmap.png"), res=350, width=9, height=5.5, units="in")
association_output$heatmap
dev.off()
```

### **Fix batch effects using SVA**

```{r}
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
mdata <- colData(dds) %>% data.frame()
svseq <- readRDS(here("QC", "results", str_interp("sva_15SVs.rds")))
SVs <- svseq$sv
colnames(SVs) <- paste0("SV", 1:15)
tmp_mdata <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)

# use limmas removeBatchEffects to emulate what the linear model will eventually do.
norm_data <- assay(vst(dds, blind=TRUE))
mod <- model.matrix(~group, tmp_mdata)
# Create the formula dynamically
formula <- as.formula(paste("~", paste0("SV", 1:15, collapse = " + ")))
mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

# adjust the counts using limma
adjusted_counts <- limma::removeBatchEffect(norm_data,
    covariates = mod0,
    design = mod
)

# recalculate PCA with adjusted counts
pca_data <- PCAtools::pca(adjusted_counts,
    metadata = colData(dds),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))
for (k in 1:20) {
    pca_plot_df[, str_interp("PC${k}_explained_var")] <- pca_data$variance[k]
}


############
# SV PLOTS #
############
# want to highlight SV3 and SV4 which nicely account for sex
SV_pairings <- lapply(1:2, function(x) {
    c(2 * x - 1, 2 * x)
})

for (k in SV_pairings) {
    print(str_interp("Working on SV${k[1]} and SV${k[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        p <- density_scatter(
            df = tmp_mdata,
            x_variable = str_interp("SV${k[1]}"),
            y_variable = str_interp("SV${k[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "ppt_scripts", "imgs",
            str_interp("SV_all_covariates_SV${k[1]}_SV${k[2]}_SVA15_corrected.png")
        ),
        dpi = 350,
        width = 18,
        height = 11.5,
    )
}


#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
# create PC intervals
PC_pairings <- lapply(1:3, function(x) {
    c(2 * x - 1, 2 * x)
})

for (k in PC_pairings) {
    print(str_interp("Working on PC${k[1]} and PC${k[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${k[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${k[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${k[1]}"),
            y_variable = str_interp("PC${k[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${k[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${k[2]}: ${expl_var2}% variance")
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "ppt_scripts", "imgs",
            str_interp("PCA_all_covariates_PC${k[1]}_PC${k[2]}_SVA15_corrected.png")
        ),
        dpi = 350,
        width = 18,
        height = 11.5,
    )
}


################
# BOXPLOTS PCA #
################
p <- plot_categorical_covar(
    df = pca_plot_df,
    covar = "line",
    PC = paste0("PC", 1:4)
) %>% wrap_plots(., ncol = 2) &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(size = 5, face = "bold")
    )

p <- p +
    plot_annotation(
        title = "Cell line",
        theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
    )


ggsave(
    plot = p,
    filename = here("ppt_scripts", "imgs", "PCA_line_boxplots_SV15_corrected.png"),
    dpi = 350,
    width = 19,
    height = 9,
)


###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
    components = paste0("PC", 1:15)
)

png(here("ppt_scripts", "imgs", "association_heatmap_SVA15_corrected.png"),
    res = 350, width = 9, height = 5.5, units = "in"
)
association_output$heatmap
dev.off()



########
# UMAP #
########1
umap_data <- umap::umap(t(adjusted_counts))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>%
    left_join(., mdata)

 
################
# ORA ANALYSIS #
################
library(clusterProfiler)
hs_msigdb_df <- msigdbr::msigdbr(species = "Homo sapiens")

gene_sets <- hs_msigdb_df %>%
    filter(gs_subcat == "GO:BP" | gs_cat == "H" | gs_subcat == "CP:KEGG") %>%
    select(gs_name, gene_symbol)

# Extract the genes which most influence PC1 and PC2 and see if they are enriched in any biological domain:
for (z in 1:2) {
    print(paste("Running ORA for PC", z))
    # extract top 300 genes which most influence PC
    top_loadings_gene_names <- pca_data$xvars[order(abs(pca_data$loadings[, paste0("PC", z)]), decreasing = T)[1:300]]

    enrichment_result <- enricher(
        gene = top_loadings_gene_names,
        universe = rownames(dds),
        minGSSize = 5,
        maxGSSize = 300,
        TERM2GENE = gene_sets
    )

    # for each of the biological domains, grep the top 3 sig. terms
    # for each of the biological domains, grep the top 3 sig. terms
    sig_results <- enrichment_result@result %>%
        mutate(domain = str_extract(Description, "^[^_]*")) %>%
        mutate(ratio_numeric = sapply(.$GeneRatio, function(x) {
            eval(
                str2expression(x)
            )
        })) %>%
        filter(p.adjust < 0.05) %>%
        group_by(domain) %>%
        slice_min(., order_by = p.adjust, n = 3) %>%
        mutate(Description_short = shorten_strings(Description, 30))

    if (nrow(sig_results) > 0) {
        enrichment_plot <- ggplot(sig_results, aes(x = ratio_numeric, y = Description_short, color = p.adjust)) +
            geom_segment(aes(
                x = 0, xend = ratio_numeric, y = Description_short, yend = Description_short
            ), color = "black", linewidth = 1.5) +
            geom_point(size = 5) +
            scale_color_gradient(low = "red", high = "blue") +
            theme_Publication_side() +
            ylab("Description") +
            xlab("Gene Ratio") +
            scale_x_continuous(expand = c(0, 0.01))
    } else {
        enrichment_plot <- ggplot() +
            theme_void()
    }
    ggsave(
        plot = enrichment_plot,
        file = here(
            "ppt_scripts", "imgs",
            str_interp("ORA_PC${z}.png")
        ),
        dpi = 150,
        width = 10,
        height = 6
    )
}
```


### Fix batch effects using donor 
The experiment was randomized in a way, that all subject are processed in the same batch (i.e. donor and batch are confounded).
This means that by adjusting for donor, we are also capturing batch effects. And because we 
are comparing samples from the same donor to one another (paired setup), sex, age, ethnicity will also be corrected. 

```{r}
# this object has the tiny libsize outliers removed
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
mdata <- colData(dds) %>% data.frame()
mod <- model.matrix(~group, colData(dds))
norm_cnts <- assay(vst(dds))
adjusted_counts <- limma::removeBatchEffect(norm_cnts,
    batch = mdata$line, 
    design = mod
)

#################
# CALCULATE PCA #
#################
pca_data <- PCAtools::pca(adjusted_counts,
    metadata = colData(dds),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))
for (i in 1:20) {
    pca_plot_df[, str_interp("PC${i}_explained_var")] <- pca_data$variance[i]
}

columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")

PC_pairings <- lapply(1:3, function(x){
    c(2*x-1,2*x)
})
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
for (i in PC_pairings) {
    print(str_interp("Working on PC${i[1]} and PC${i[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${i[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${i[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${i[1]}"),
            y_variable = str_interp("PC${i[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
        )


        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "ppt_scripts", "imgs",str_interp("PCA_covariates_PC${i[1]}_PC${i[2]}_donor_corrected.png")
        ),
        dpi = 350,
        width = 18,
        height = 11.5,
    )
}

################
# BOXPLOTS PCA #
################
p <- plot_categorical_covar(
    df = pca_plot_df,
    covar = "batch1id",
    PC = paste0("PC", 1:4)
) %>% wrap_plots(., ncol = 2) &
    theme(
        axis.text.x = element_blank(),
        axis.text.y = element_text(size=5, face="bold")
    )
p <- p +
    plot_annotation(
        title = "Batch",
        theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
    )

ggsave(
    plot = p,
    filename = here("ppt_scripts", "imgs", "PCA_batch_boxplots_donor_corrected.png"),
    dpi = 350,
    width = 10,
    height = 9,
)

#########
#  UMAP #
#########
umap_data <- umap(t(adjusted_counts))

umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>%
    left_join(., colData(dds) %>% data.frame())

columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
ggsave(
    plot = final_plot,
    filename = here(
        "ppt_scripts", "imgs", "UMAP_covariates_donor_corrected.png"
    ),
    dpi = 350,
    width = 18,
    height = 11.5,
)
```


### Fix batch effects using combination of donor and SVA

```{r}
# load in DESeq2dataset where the libsize outliers are already removed.
# for creation see script **01_initial_QC.rmd**
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
dds <- estimateSizeFactors(dds)
norm_counts <- counts(dds, norm=TRUE)
sva_seq <- readRDS(here("QC", "results", "sva_incl_line_5SVs.rds"))
SVs <- sva_seq$sv
colnames(SVs) <- paste0("SV", 1:5)
tmp_mdata <- cbind(colData(dds) %>% data.frame() %>% select(-contains("SV")), SVs)

# use limmas removeBatchEffects to emulate what the linear model will eventually do.
norm_data <- assay(vst(dds, blind=TRUE))
mod <- model.matrix(~group, tmp_mdata)
# Create the formula dynamically
formula <- as.formula(paste("~", paste0("SV", 1:5, collapse = " + ")))
mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

# adjust the counts using limma
adjusted_counts <- limma::removeBatchEffect(norm_data,
    batch = dds$line,
    covariates = mod0,
    design = mod
)

# recalculate PCA with adjusted counts
pca_data <- PCAtools::pca(adjusted_counts,
    metadata = colData(dds),
    removeVar = 0.4
)
pca_plot_df <- pca_data$rotated[, 1:20] %>%
        rownames_to_column("sample_id") %>%
        left_join(., mdata, by = c("sample_id" = "sampleid"))
for (k in 1:6) {
    pca_plot_df[, str_interp("PC${k}_explained_var")] <- pca_data$variance[k]
}

#############
# PCA PLOTS #
#############
columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
# create PC intervals
PC_pairings <- lapply(1:3, function(x) {
    c(2 * x - 1, 2 * x)
})

for (k in PC_pairings) {
    print(str_interp("Working on PC${k[1]} and PC${k[2]}"))
    plots <- lapply(columns_to_plot, function(x) {
        expl_var1 <- round(pca_plot_df[1, str_interp("PC${k[1]}_explained_var")], 2)
        expl_var2 <- round(pca_plot_df[1, str_interp("PC${k[2]}_explained_var")], 2)
        p <- density_scatter(
            df = pca_plot_df,
            x_variable = str_interp("PC${k[1]}"),
            y_variable = str_interp("PC${k[2]}"),
            color_by = x,
            density_plot_ratio = 0.15,
            pt_size = 1.2,
            xlab = str_interp("PC${k[1]}: ${expl_var1}% variance"),
            ylab = str_interp("PC${k[2]}: ${expl_var2}% variance")
        )
        p <- p +
            plot_annotation(
                title = x,
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
            )
        return(p)
    })
    final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
    ggsave(
        plot = final_plot,
        filename = here(
            "ppt_scripts", "imgs",
            str_interp("PCA_all_covariates_PC${k[1]}_PC${k[2]}_SVA5_incl_line_corrected.png")
        ),
        dpi = 350,
        width = 18,
        height = 11.5,
    )
}


################
# BOXPLOTS PCA #
################

# -> no associations
p <- plot_categorical_covar(
    df = pca_plot_df,
    covar = "line",
    PC = paste0("PC", 1:4)
) %>% wrap_plots(., ncol = 2) &
    theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(size = 5, face = "bold")
    )

p <- p +
    plot_annotation(
        title = "Cell line",
        theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
    )
p




###########
# HEATMAP #
###########
association_output <- covar_correlation(
    df = pca_plot_df,
    covars = c("pool", "state", "condition", "oxldlset", "sex", "ethnicity", "age", "batch1id", "libsize"),
    components = paste0("PC", 1:15)
)

png(here("ppt_scripts", "imgs", "association_heatmap_SVA5_incl_line_corrected.png"),
    res = 350, width = 9, height = 5.5, units = "in"
)
association_output$heatmap
dev.off()



########
# UMAP #
########
umap_data <- umap::umap(t(adjusted_counts))
umap_plot_df <- umap_data$layout %>%
    data.frame() %>%
    magrittr::set_colnames(c("UMAP1", "UMAP2")) %>%
    rownames_to_column("sampleid") %>%
    left_join(., mdata)

columns_to_plot <- c("state", "condition", "sex", "ethnicity", "age", "batch1id")
plots <- lapply(columns_to_plot, function(x) {
    p <- density_scatter(
        df = umap_plot_df,
        x_variable = "UMAP1",
        y_variable = "UMAP2",
        color_by = x,
        density_plot_ratio = 0.15,
        pt_size = 1.2,
    )
    p <- p +
        plot_annotation(
            title = x,
            theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 20))
        )
    return(p)
})

final_plot <- cowplot::plot_grid(plotlist = plots, ncol = 3)
ggsave(
    plot = final_plot,
    filename = here(
        "ppt_scripts", "imgs", "UMAP_covariates_SVA5_incl_line_corrected.png"
    ),
    dpi = 350,
    width = 18,
    height = 11.5,
)
```


#### Outlier detection
```{r}
dds <- readRDS(here("QC", "results", "dds_libsize_outliers_removed.rds"))
mdata <- colData(dds) %>% data.frame()
sva_seq <- readRDS(here("QC", "results", "sva_incl_line_5SVs.rds"))
dds <- estimateSizeFactors(dds)
tmp_mdata <- cbind(mdata, sva_seq$sv %>% data.frame() %>% magrittr::set_colnames(paste0("SV", 1:5)))
norm_data <- assay(vst(dds, blind = TRUE))

mod <- model.matrix(~group, tmp_mdata)
formula <- as.formula(paste("~", paste0("SV", 1:5, collapse = " + ")))
mod0 <- model.matrix(formula, tmp_mdata)[, -1] # this removes the intercept term. See this post: https://support.bioconductor.org/p/9144613/

# adjust the counts using limma
adjusted_counts <- limma::removeBatchEffect(norm_data,
    batch = dds$line,
    covariates = mod0,
    design = mod
)
mdata <- mdata %>%
    mutate(group = paste(state, condition, sep="_")) 
cor_result <- lapply(unique(mdata$group), function(x) {
    samples <- mdata %>%
        filter(group == x) %>%
        pull(sampleid)

    tmp_cnts <- adjusted_counts[, samples]
    cor_map <- cor(tmp_cnts, method = "spearman")
    return(cor_map)
})
names(cor_result) <- unique(mdata$group)
```


##### M0_ctrl
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M0_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M0_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:5) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
M0_ctrl_plot <- density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)



# MM_oxLDL9752143 is very different from the rest... remove it
M0_ctrl_outlier <- c("MM_oxLDL9752143")

```


##### M1_ctrl
```{r}

# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M1_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M1_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
M1_ctrl_plot <- density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)

# MM_oxLDL9752115 is very different from the rest, prob. because of very low libsize 
M1_ctrl_outlier <- c("MM_oxLDL9752115")

```


##### M2_ctrl
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M2_ctrl)],
    metadata = colData(dds)[rownames(cor_result$M2_ctrl),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
M2_ctrl_plot <- density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)
# low libsize again. 
M2_ctrl_outlier <- c("MM_oxLDL9752225")

```


##### M0_stim
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M0_stim)],
    metadata = colData(dds)[rownames(cor_result$M0_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
M0_oxldl_plot <- density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)


# MM_oxLDL7992361
mdata %>% filter(sampleid=="MM_oxLDL7992361")
M0_oxldl_outlier <- c("MM_oxLDL7992361")
```


##### M1_stim
```{r}
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M1_stim)],
    metadata = colData(dds)[rownames(cor_result$M1_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}
i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
M1_oxldl_plot <- density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)


M1_oxldl_outlier <- "MM_oxLDL7734278"
```


##### M2_stim
```{r}
# This looks just fine. No outliers
# recalculate PCA with adjusted counts
pca_data_subset <- PCAtools::pca(adjusted_counts[, rownames(cor_result$M2_stim)],
    metadata = colData(dds)[rownames(cor_result$M2_stim),],
    removeVar = 0.4
)
pca_plot_df_subset <- pca_data_subset$rotated[, 1:20] %>%
    rownames_to_column("sample_id") %>%
    left_join(., mdata, by = c("sample_id" = "sampleid"))

for (i in 1:20) {
    pca_plot_df_subset[, str_interp("PC${i}_explained_var")] <- pca_data_subset$variance[i]
}

i = c(1, 2)
expl_var1 <- round(pca_plot_df_subset[1, str_interp("PC${i[1]}_explained_var")], 2)
expl_var2 <- round(pca_plot_df_subset[1, str_interp("PC${i[2]}_explained_var")], 2)
M2_oxldl_plot <- density_scatter(
    df = pca_plot_df_subset %>% mutate(`libsize ` = ifelse(libsize < 175000, "<175000", ">175000")),
    x_variable = str_interp("PC${i[1]}"),
    y_variable = str_interp("PC${i[2]}"),
    color_by = "libsize ",
    density_plot_ratio = 0,
    pt_size = 3,
    xlab = str_interp("PC${i[1]}: ${expl_var1}% variance"),
    ylab = str_interp("PC${i[2]}: ${expl_var2}% variance")
)


```


##### Combine plots
```{r}
final_plot <- wrap_plots(
    M0_ctrl_plot, M1_ctrl_plot, M2_ctrl_plot,
    plot_spacer(), plot_spacer(), plot_spacer(),
    M0_oxldl_plot, M1_oxldl_plot, M2_oxldl_plot
) + plot_layout(heights = c(1, 0.15, 1))


ggsave(
    plot = final_plot,
    filename = here(
        "ppt_scripts", "imgs", "group_wise_PCAs_corrected.png"
    ),
    dpi = 350,
    width = 16,
    height = 12.5,
)
```


#### Plot PCA and UMAP after removing the samples. 


```{r}
sessioninfo()
```