---
title: "GeneID to GeneName mapping" 
author: "Nick Hirschmuller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  html_document:
    word_document:
    toc: yes
    toc_depth: '3'
    code_folding: hide
  pdf_document:
    number_sections: yes
    toc: false
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = T, warning = F, error = F, message = F,
    fig.align = "center", cache = F, dpi = 350
)
```

## **Introduction**
The output of the [nf-core RNAseq pipeline](https://github.com/nf-core/rnaseq) already contains count tables. 
However, in this table there are some ensembl gene names that map to multiple ensembl IDs. 
This will create problems downstream because we want to primarily use the gene names.
In this script we will make sure that the mapping is 1 to 1. 

```{r, load libraries}
library(tidyverse)
library(here)
```


See nf-core rna seq documentation:
https://nf-co.re/rnaseq/3.10.1/docs/output#salmon -> salmon.merged.gene_counts_length_scaled use this directly as you would with a regular counts matrix.
Also see comment from Mike love here: https://support.bioconductor.org/p/9149309/
He recommends to use salmon.merged.gene_counts_length_scaled.tsv as the count matrix input.

```{r load the count table of interest}
raw_counts <- data.table::fread(here("data", "oxLDL_data", "RNA_seq", "results", "star_salmon", "salmon.merged.gene_counts_length_scaled.tsv"),
    data.table = FALSE
)
gene_ids <- raw_counts %>%
    group_by(gene_name) %>%
    mutate(n_ = n()) %>%
    filter(n_ > 1) %>%
    pull(gene_id) 

subset_counts <- raw_counts %>%
    filter(gene_id %in% gene_ids) %>% 
    arrange(gene_name)

# we will keep the gene_id which has more counts on average across all samples
subset_counts$MM_avg <- rowMeans(subset_counts[, grepl("^MM", names(subset_counts))])
keep_gene_ids <- subset_counts %>%
    group_by(gene_name) %>%
    slice_max(MM_avg, n = 1, with_ties = FALSE) %>%
    pull(gene_id)
remove_gene_ids <- setdiff(gene_ids, keep_gene_ids)

raw_counts_filtered <- raw_counts %>%
    filter(!gene_id %in% remove_gene_ids)

# QC check -> should be empty.
raw_counts_filtered %>%
    group_by(gene_name) %>%
    mutate(n_ = n()) %>%
    filter(n_ > 1) %>%
    pull(gene_id) 

data.table::fwrite(
    x = raw_counts_filtered %>% select(-gene_id),
    file = here("preprocessing", "results", "gene_counts_length_scaled_filtered.tsv"),
    sep = "\t", row.names = FALSE
)


```